Loose coupling between Ca2+ channels and release sensors as a synaptic correlate of higher order brain function

Version 1

Abstract

In the mature neocortex, functionally distinct areas are built by the same archetypes of neurons, but depending on the area, these neurons and their synapses are engaged in very different functions, ranging from lower order processing of sensory information to higher order associations and cognitive functions. We found significant differences in the functional presynaptic nanoarchitectures of the same types of pyramidal neuron synapses, depending on whether they are located in the prefrontal cortex (PFC) or in the primary somatosensory cortex (S1). Synapses in PFC operated with loose microdomain coupling as opposed to tight nanodomain coupling in S1. These differences were associated with significant differences in synaptic timing, efficacy and plasticity between areas. Our data suggest that the mature neocortex uses tuning of synaptic topographies to specialize seemingly identical types of neurons for their required function. They reveal microdomain coupling as a presynaptic structure-function correlate of higher order neocortical functions.

Introduction

The neo- or isocortex is a morphologically homogeneous brain region that covers highly diverse functions within its specialized areas, ranging from early sensory processing and motor control up to higher order associations, cognition and consciousness (Douglas and Martin, 2004; Harris and Shepherd, 2015). Notably, all of these diverse functions are executed by the same archetypes of neurons and synapses. However, depending on the area, they are engaged to different degrees in lower or higher order processing. Whether area-specific functional differences in the mature neocortex are associated with or even arise from differences in the functional presynaptic nanoarchitecture of the same principal types of synapses is currently unclear.

At presynaptic active zones, the coupling distance between voltage-gated Ca2+ channels (VGCCs) and transmitter-filled synaptic vesicles (SVs) is a major determinant of key synaptic properties, including speed, efficacy and reliability of synaptic transmission (Rozov et al., 2001; Fedchyshyn and Wang, 2005; Bucurenciu et al., 2008; Baur et al., 2015; Bornschein et al., 2019a; Chen et al., 2024), and also short-term plasticity, although the latter can be obscured by vesicle replenishment (Miki et al., 2016; Doussau et al., 2017; Bornschein et al., 2019b; Lin et al., 2022). Studies at different synapses in different parts of the brain suggest that excitatory synapses engaged in reliable information transfer switch from loose microdomain to tight nanodomain coupling during development (Fedchyshyn and Wang, 2005; Baur et al., 2015; Nakamura et al., 2015; Bornschein et al., 2019a). This includes synapses between layer 5 pyramidal neurons (L5PNs) in the primary somatosensory cortex (S1), an area of early sensory processing (Bornschein et al., 2019a). On the other hand, loose microdomain coupling was found in the mature brain to date only at a highly plastic hippocampal synapse and has been suggested to provide a molecular framework for presynaptic plasticity (Vyleta and Jonas, 2014). However, if microdomain coupling also represents a synaptic correlate of higher order neocortical function or if it restricted to specific and highly plastic synapses in the hippocampus is currently unclear. Therefore, here we tested the hypothesis that loose microdomain coupling is the synaptic correlate of higher order functions of the mature neocortex, whereas tight nanodomain coupling is the synaptic correlate of earlier processing stages. Our data show that in the mature cortex the same types of pyramidal neuron synapses operate with loose microdomain coupling if they are located in the prefrontal cortex (PFC), whereas in S1 they operate with tight nanodomain coupling. Thus, our results corroborate the hypothesis and suggest microdomain coupling as a synaptic correlate of higher order function in the neocortex.

Results

Synaptic efficacy, reliability and short-term plasticity differ between PFC and S1

We focused on the lateral prefrontal cortex (PFC) and medial prefrontal cortex (mPFC) as areas of higher order processing and on primary somatosensory cortex (S1) as an area of lower order processing. Within these areas we investigated excitatory synapses between pyramidal neurons (PNs), which are the principal building blocks of the neocortex. We first analyzed the properties of synapses connecting neighboring layer 5 PNs (L5PNs) in mature PFC (P21-26) by performing paired whole-cell patch-clamp recordings and compared their properties to those we previously reported for the same synapses and age in S1 (Bornschein et al., 2019b; Bornschein et al., 2019a) (Figure 1A). We found that the efficacy of transmission, i.e. the size of a unitary EPSC evoked by a single action potential (Rozov et al., 2001), was significantly lower in PFC (8 pA, 5-15 pA) than in S1 (29 pA, 18-53 pA), whereas the synaptic failure rates were significantly higher in PFC (0.16, 0.08-0.28) than in S1 (0, 0-0.03; Figure 1B-D). In addition, we found that the synaptic delays (2.46 ms, 2.20-2.94 ms) as well as their standard deviation (SDDelay: 0.60 ms, 0.49-0.81 ms) were significantly larger in PFC than in S1 (delay: 1.99 ms, 1.67-2.25 ms; SDDelay: 0.24 ms, 1.14-0.37 ms; Figure 1E, F), indicating that transmitter release is less tightly coupled to the time of the action potential in PFC than in S1 (Bullmann et al., 2024). Together these data show that synaptic speed, efficacy and reliability during single action potentials are lower in PFC than at the same synapses in S1.

Differences in synaptic transmission and plasticity between PFC and S1 synapses.

(A) Schematic representation of the mature mouse brain (top) and the positions at which coronal sections were made from PFC (orange; middle) or S1 (blue; bottom). Recordings were made in acute slices from L5PN-L5PN pairs or from L2/3-L5PN connections. (B) EPSCs (bottom) recorded from postsynaptic L5PNs after evoking action currents in the presynaptic L5PNs (top, inset: 3fold magnification) in PFC (left) or S1 (right), individual recordings in gray, averages in black). (C-F) Summary of basic synaptic properties: EPSC amplitudes (C), failure rates (D), synaptic delays (E) and their SD values (F) in PFC and S1 (medians ± IQRs, means as dashed lines, whiskers represent 10th and 90th percentiles, dots indicate outliers, numbers of cell pairs in brackets; ***P<0.001, MWU). Data from S1 are from ref. (Bornschein et al., 2019a). (G) EPSCs recorded from pairs of L5PNs at 50 Hz intervals in PFC (top) and S1 (bottom), presynaptic stimulations are omitted, averages from at least 10 individual recordings). (H, I) Summary of PPRs (H; ***P<0.001, MWU) and success rates (I) following the 1st (1_x) and 2nd stimulation (x_1; ***P<0.001, P=0.987, WSR). S1 data are from ref. (Bornschein et al., 2019a). (J) As in (G) but for L2/3-L5 synapses (bottom, averages from at least 15 individual recordings, stimulation artefacts were blanked for clarity, arrowheads denote the time points of extracellular stimulation). (K) Boxplots summarizing PPR data from lateral PFC (PFC), medial PFC (mPFC), older PFC (P90-100) and in S1 (***P<0.001, MWU; n.s., not significant, P=0.067, ANOVA on ranks). (L) Comparison of SDs of the synaptic delays at L2/3-L5 synapses between PFC and S1 (***P<0.001, MWU).

When pairs of action potentials were applied at high frequency (Figure 1G-I), differences in the synaptic success rates were associated with a significant difference in the paired pulse ratios (PPR) between PFC and S1. In PFC paired pulse facilitation (PPF, 1.18, 0.98-1.52) prevailed associated with increased success rates in the 2nd pulse (1_x: 0.81, 0.73-0.91; x_1: 0.91, 0.82-0.97). In contrast, in S1 paired pulse depression (PPD, 0.78, 0.69-0.90) was found with high initial success rates (1_x: 1.00, 0.97-1.00; x_1: 0.99, 0.95-1.00). We observed the difference in short-term plasticity over a range of interstimulus intervals (ISIs) of 5 to 50 ms (SI Appendix, Figure S1A, B). Thus, the differences in synaptic efficacy and reliability between S1 and PFC during the first action potential correlate with differential short-term plasticity, with the emphasize in S1 being on the first transmission process while in PFC facilitation emphasized the subsequent release.

Next, we investigated transmission at synapses between L2/3 and L5PNs, using minimal extracellular stimulation near somata of L2/3PNs (Figure 1A). We found that also L2/3-L5PN synapses in PFC showed PPF (1.36, 1.14-1.88) as opposed to PPD in S1 (0.87, 0.72-0.99; Figure 1J, K). We observed this over a range of ISIs up to 100 ms (SI Appendix, Figure S1C, D). In addition, at PFC synapses the SD of the synaptic delays (0.46, 0.29-0.66) was again significantly larger than in S1 (0.25, 0.15-0.37; Figure 1L). Furthermore, we found larger EPSC decay time constants (τdecay) in PFC compared to S1, possibly reflecting a contribution of different postsynaptic glutamate receptors (SI Appendix, Figure S2A, B). Indeed, application of the N-methyl-D-aspartate receptor (NMDAR) antagonist APV removed the difference in τdecay between PFC and S1 (SI Appendix, Figure S2C-F). We extended the paired pulse experiments at L2/3-L5PN synapses in PFC to even older mice (P90-100) and to the mPFC (P21-26). We found that PPF persisted in the older age window (1.25, 1.14-1.41) and that also the synapses in mPFC showed PPF (1.14, 1.03-1.42; Figure 1K).

Together these results from glutamatergic synapses onto L5PNs show significant differences in the synaptic delays, efficacy, reliability and short-term plasticity between PFC and S1. The results would be consistent with area-specific differences in the functional coupling distances between VGCCs and SVs at the presynaptic active zones.

EGTA sensitivity of release is higher in PFC than in S1

To directly test for differences in the coupling distances, we investigated the sensitivity of release to synthetic Ca2+ chelators (Figure 2). The sensitivity of release to low millimolar concentrations of the kinetically slow Ca2+ chelator EGTA is a standard indicator of loose coupling (Eggermann et al., 2012). First, we performed recordings at pairs of L5PNs in mature PFC (Figure 2A, B). After 10 min of stable baseline recordings, the presynaptic neuron was patched a second time and equilibrated for 30 min with 10 mM EGTA-containing pipette solution. In these paired recordings we found a significant effect of EGTA onto release in PFC (0.67, 0.63-0.73; control: 1.02, 0.75-1.20). In identical experiments in S1, we had previously found release to be insensitive to the same concentration of EGTA (0.94, 0.71-1.02) (Bornschein et al., 2019a). These data provide evidence that the synapses between L5PNs in mature PFC do indeed operate with loose coupling, in contrast to their counterparts in S1.

Area-specific differences in EGTA sensitivity.

(A) Averaged, baseline-normalized EPSC amplitudes (means ± SEMs, 2-min bins) recorded from L5PN-L5PN pairs in PFC with normal pipette solution (control, orange) and after re-patching the presynaptic neuron with a pipette solution containing 10 mM EGTA (blue). Insets: Average EPSCs from one pair during baseline (-10 to 0 min) and test period (20 to 30 min) for control and EGTA (averages normalized to baseline). (B) Boxplot comparing the normalized EPSC amplitudes prior and following the application of EGTA with control recordings in L5PN pairs in PFC (*P=0.041, MWU). Note the significant EGTA effect. (C) As in (A) but for L2/3-L5PN connections in PFC. Following baseline recordings (10 min), slices were perfused for 30 min (solid line) with ACSF containing either 10 μM EGTA-AM (blue), 10 μM BAPTA-AM (red), or only the solvent DMSO/Pluronic (control, orange), and thereafter rinsed for 10 min with ACSF (test period). Insets: Averaged EPSCs during baseline and test period. (D) Summary of chelator-AM effects in PFC, mPFC and older PFC (P90-100; *P=0.016, **P=0.007, 0.008, 0.008, MWU; P=0.404, ANOVA on ranks). Note the EGTA sensitivity of release in all PFC recordings. (E) As in (C) but for L2/3-L5PN connections in S1 (control in light blue). (F) Summary of chelator-AM effects in S1 (P=0.792, **P=0.008, MWU). Note the absence of a significant EGTA effect in S1.

We proceeded by comparing the EGTA sensitivity of release at L2/3-L5PN synapses between S1 and PFC, using extracellular stimulation and incubation with Ca2+ chelator-AM compounds (Figure 2C-F) (Baur et al., 2015; Kusch et al., 2018). Following 10 min of stable baseline recordings, slices were perfused with ACSF that either contained the DMSO/Pluronic containing solvent alone (control) or 10 µM of dissolved EGTA-AM or BAPTA-AM. The fast BAPTA interferes with release irrespective of the coupling topography, thereby, providing a positive control for the proper functioning of the AM-method (Eggermann et al., 2012). EPSC amplitudes were quantified during a subsequent 10 min period with perfusion with normal ACSF. During this test period, BAPTA significantly reduced EPSC amplitudes both in PFC (0.28, 0.12-0.32; control: 0.99, 0.84-1.02) and in S1 (0.08, 0-0.18; control: 0.85, 0.83-0.92). On the other hand, EGTA significantly reduced EPSC amplitudes only in PFC (0.57, 0.50-0.70; 58% of control) but not in S1 (0.85, 0.81-0.90; 100% of control).

In S1 we previously found that coupling switches from loose to tight during postnatal development between ∼P10 and ∼P20 (Bornschein et al., 2019a). To probe for a developmental retardation of this process in PFC, we extended the EGTA-AM experiments at L2/3-L5PN synapses to the older age-window of P90-100. Also, in these experiments we found a significant sensitivity of release to EGTA (0.76, 0.65-0.78; Figure 2D; SI Appendix, Figure S3A), indicating that loose coupling is a functionally persistent feature of synapses in PFC. Finally, we also extended the EGTA-AM experiments to synapses in mPFC and again found significant effects of EGTA on the EPSC amplitudes (0.65, 0.49-0.75; Figure 2D; SI Appendix, Figure S3B).

Together the results obtained with the slow Ca2+ chelator EGTA suggest that major glutamatergic synapses in mature PFC operate with loose coupling in contrast to tight coupling in the mature S1. The data further suggest that loose coupling is a developmentally persistent property of synapses in PFC. Finally, they indicate that the above functional differences between synapses in S1 and PFC originate from the differences in Ca2+ influx-release coupling.

Release probabilities are similar in PFC and S1

To test whether the release probabilities (pN) of the differentially coupled release sites in PFC and S1 differ, we performed multiple probability fluctuation analysis (MPFA) (Clements and Silver, 2000; Brachtendorf et al., 2025) at different extracellular Ca2+ concentrations ([Ca2+]e; Figure 3) (Bornschein et al., 2019a). The parabolic MPFA-fits to variance-mean plots from recordings of pairs of L5PNs in PFC yielded a median pN of 0.50 (0.37-0.55; Figure 3A, D). This value was slightly but not significantly smaller than pN of 0.66 (0.55-0.70), which we quantified previously for this connection in S1 (Bornschein et al., 2019a).

Quantal synaptic parameters in PFC and S1.

(A) MPFA of EPSC amplitudes recorded at the indicated [Ca2+]e from a pair of L5PNs in PFC. Top: EPSCs (gray, average in black) recorded in 1, 2 and 5 mM [Ca2+]e. Middle: Plot of EPSC amplitudes over time. Bottom: Corresponding mean-variance plot fitted with a parabola estimating the quantal parameters of release. pN is for 2 mM [Ca2+]e. (B) As in (A), but for a L2/3-L5PN connection in PFC. Recordings were made in the presence of 0.25 mM Kyn and 50 µM APV. (C) As in (B), but in S1. (D) Summary of quantal release parameters in L5PN pairs from PFC and S1 and in S1 with 1-2 mM γDGG (cf. (1); P=0.121, *P=0.048, ANOVA on ranks; P=0.054, MWU PFC vs. S1). (E) Left: Example of evoked qEPSCs (top) and the corresponding amplitude histogram (bottom, gray range was used for calculating the qEPSC amplitude). Right: Summary of qEPSC amplitudes in PFC and S1 (**P=0.004, MWU). (F) Summary of pN values in L2/3-L5PN connections from PFC and S1 (P=0.222, MWU).

We also quantified pN at L2/3-L5PN synapses, both in PFC and in S1, using MPFA with extracellular stimulation (Baur et al., 2015). We found that pN was as high as at the L5PN-L5PN synapses in both areas with values (PFC: 0.62, 0.47-0.67; S1: 0.48, 0.39-0.58) being not significantly different between areas (Figure 3B, C, F). Interestingly, in our previous study (Bornschein et al., 2019a) we found that the developmental switch from loose to tight coupling at L5PN synapses in S1 also had no significant impact on pN.

From MPFA in paired recordings it is possible to estimate the number of release sites (N) and the quantal size (q), additionally. N was significantly smaller in PFC than in S1 being only 2.1 (1.9-2.3) rather than 8 (3-19). q tended to be smaller in PFC (6 pA, 5-7 pA) than in S1 (9 pA; 8-11 pA) and the difference became significant when evoked quantal EPSCs recorded in low [Ca2+]e were compared (PFC: 4 pA, 4-5 pA; S1: 9 pA, 7-11 pA). We consider it unlikely that postsynaptic receptor saturation or desensitization influenced the determination of quantal parameters since experiments with the competitive low-affinity glutamate receptor antagonist γ-DGG (γ-D-glutamylglycine) (Chanda and Xu-Friedman, 2010) performed at the S1 synapses had revealed no signs of this (pN=0.58, 0.52-0.60; N=8, 8-12; Figure 3D). Due to the smaller size of the EPSCs, testing on the PFC synapses was not possible. Estimating EPSC amplitudes (EPSC = N pN q; PFC, 6 pA; S1, 48 pA) and failure rates (F = (1-pN)^N; PFC, 0.25, S1, 0.0001) from the quantal parameters yielded values that were well suited to explain the area-specific differences in synaptic efficacy during individual action potentials (Figure 1).

Presynaptic Ca2+ signals are similar in PFC and S1

For a thorough interpretation of the above results knowledge about the presynaptic Ca2+ dynamics in the different synapses is required. We performed dual-dye two-photon Ca2+ imaging (Sabatini et al., 2002) at presumed presynaptic boutons located on axon collaterals of L5PNs in PFC and in S1. We quantified single action potential-mediated elevations in green over red fluorescence signals (ΔG/R) that were converted to increases in intracellular calcium (Δ[Ca2+]i) (Figure 4) (Bornschein et al., 2019a). We found that neither the Ca2+ transients nor the decay time constants were significantly different between the cortical areas (Figure 4A-F; PFC: G/R=0.25, 0.23-0.31; Δ[Ca2+]i=279 nM, 244-453 nM; basal [Ca2+]i=30 nM, 21-34 nM; τdecay,1=9 ms, 6-19 ms; τdeay,2=91 ms, 75-120 ms; S1: G/R=0.29, 0.25-0.39; Δ[Ca2+]i=238 nM, 208-303 nM; basal [Ca2+]i=22 nM, 13-38 nM; τdecay,1=9 ms, 3-15 ms; τdecay,2=86 ms, 63-212 ms). The similarity in the presynaptic Ca2+ transients of boutons in PFC and S1 suggests that differences in the presynaptic Ca2+ influx are unlikely to be the cause for the differential Ca2+ chelator effects. Rather, they support the view that these differences result from loose coupling in PFC and from tight coupling in S1.

Presynaptic Ca2+ transients are similar in PFC and S1.

(A) Two-photon images of L5PNs in PFC (left) and S1 (right) filled with Fluo-5F and Alexa 594 (scale bar 50 µm). Insets: Magnifications showing presynaptic boutons from which Ca2+ transients are recorded (scale bar 5 µm). (B) G/R signals (top) evoked by single APs in individual cells (gray, average of 5-13 boutons each, grand average in black) and their conversion to Ca2+ increases (Δ[Ca2+]i, bottom). (C-F) Summary of G/R signals (C; P=0.112, MWU), Δ[Ca2+]i (D; P=0.361), basal Δ[Ca2+]i (E; P=0.597) and decay time constants τ1 and τ2 (F; P=0.532, 0.962) in PFC and S1. Note, that there is no significant difference in presynaptic Ca2+ transients between PFC and S1.

Estimate of the coupling distance in PFC verify loose microdomain coupling in PFC

To derive a quantitative picture of the coupling topography in L5PN synapses in PFC we used numerical computer simulations that were based on and constrained by the experimental data. First, we simulated the measured Ca2+ transient with the indicator dye present in the simulation (Figure 5A). Subsequently, the indicator dye was removed and a release sensor model for Synaptotagmin-1-triggered release from L5PN boutons was included in the simulations (Figure 5B, C; SI Appendix, Figure S4) (Bornschein et al., 2025). VGCCs were assumed to form a ring like structure around a vesicle (Figure 5D), similar to the release site topography in young S1 (Bornschein et al., 2019a). Release rates as reported by the sensor model were integrated over time to yield the pN values. The coupling distance between the sensor and the ring of VGCCs as well as the number of VGCCs forming the ring were iteratively varied until the simulation correctly predicted the experimental pN values from MPFA obtained under control conditions and in the presence of EGTA. By this procedure we quantified an average coupling distance of 48-51 nm between a microdomain of several VGCCs and the release sensor (Figure 5B; SI Appendix, Figure S4), similar to the estimate for these synapses in immature S1 (Bornschein et al., 2019a). Thus, a microdomain model similar to immature S1 is suitable to predict the data from mature PFC, whereas the same synapses in mature S1 operated with Ca2+ signaling nanodomains in which only 1-3 VGCCs at distances of 11-16 nm trigger fusion (Bornschein et al., 2019a; Bornschein et al., 2025).

Model predicts microdomain coupling at PFC synapses.

(A) Fit of the model (blue line) to the measured average Ca2+ signal (A; gray line) of L5PN boutons in PFC. The fit was obtained with the indicated number of open VGCCs. (B) pN values were simulated with the Syt1 priming model (Bornschein et al., 2019a; Bornschein et al., 2025) for control conditions and in the presence of EGTA with different VGCC numbers yielded different coupling distances (CD) in the model. The simulated pN were subtracted from and normalized to the experimental values, i.e. a value close to zero indicates best agreement between simulation and experiment. This was obtained with 9 open VGCCs at a CD of 51 nm (see C). (C) Simulated pN for 9 open VGCCs under control conditions at increasing CDs (blue line). The experimental pN value (median ± IQR, dashed line, gray range, Figure 3D) was reproduced at a CD of 51 nm (48-55 nm). (D) Illustration of the ring model used in B and C with a CD of 51 nm between 9 VGCCs (orange) and the release sensor of a vesicle (blue). (E) Examples of averaged normalized paired pulse recordings at 50 Hz at L2/3-L5PN synapses before (black) and after (orange) application of 5 µM Cd2+ in PFC. (F) Bar graph quantifying normalized PPRs after Cd2+ application in PFC and S1 (**P=0.005, P=0.667, paired t-Test). Note that Cd2+ increased the PPR in PFC.

To experimentally test whether overlapping Ca2+ domains from several VGCCs trigger release in PFC (microdomain), we analyzed the effects of the unspecific VGCC blocker Cd2+ onto the PPR. If fusion is triggered by such VGCC microdomains, application of a subsaturating concentration of Cd2+ will increase the PPR, while it will leave PPR unaffected if only a single or few VGCCs trigger release (Hefft et al., 2002; Bucurenciu et al., 2008; Scimemi and Diamond, 2012; Baur et al., 2015; Bornschein et al., 2019a). In paired pulse experiments, at L2/3-L5PN synapses in PFC we indeed found a significant increase in the PPR following the application of 5 µM Cd2+ (rPPR=1.20±0.06), whereas the average PPR remained unaffected in S1 (rPPR=1.01±0.10; Figure 5E, F). Thus, our data and simulations suggest that the nanotopography of PN synapses in mature PFC is reminiscent of the same synapses in young S1.

Discussion

Our results provide evidence that the same archetypes of glutamatergic synapses show area-specific differences in their functional release site nanoarchitectures in the mature neocortex. They suggest loose microdomain coupling as a synaptic correlate of higher order neocortical functions.

The differences in coupling gave rise to larger synaptic delays in PFC compared to S1 and were associated with differences in synaptic reliability and efficacy, whereas the pN values were not significantly different between areas. A major determinant of pN is the size of the Ca2+ signal at the release sensor. Accordingly, tightening of coupling can increase pN (Eggermann et al., 2012; Baur et al., 2015; Bornschein and Schmidt, 2019). However, large Ca2+ elevations at the release sensor can also be obtained with Ca2+ microdomains. At L5PN-L5PN synapses in S1 the developmental switch from microdomain to nanodomain coupling had no effect on pN (Bornschein et al., 2019a). At the calyx of Held (Iwasaki and Takahashi, 2001; Taschenberger et al., 2002; Fedchyshyn and Wang, 2005; Koike-Tani et al., 2008; Nakamura et al., 2015) and at cerebellar basket cell to Purkinje cell synapses (Chen et al., 2024) pN even decreased during development, whereas coupling got tightened. The decrease in pN was compensated by an increase in N at these synapses, which increased their reliability and efficacy during development. We found N to be larger in S1 than in PFC. Thus, it appears that differences in coupling are typically associated with other characteristic changes in the release parameters. Overall, this suggests that reliable synapses switch from an initial state with high plasticity to a matured state with high reliability. On the other hand, PN synapses in the PFC appear to remain in a more juvenile state that favors plasticity over reliability.

We observed differences in short-term plasticity that suggest that S1 emphasized the first transmission process, whereas in PFC facilitation emphasized the subsequent release. The mechanisms of short-term plasticity are complex and include the regulation of pN, the speed of vesicle replenishment and the recruitment of release sites (Neher and Brose, 2018; Schmidt, 2019; Neher, 2023; Brachtendorf et al., 2025). We consider the differential emphasize on the available presynaptic efficacy between PFC and S1 as a signature of higher plasticity and flexibility in the processing of information in PFC compared to S1.

Our results have implications for long-term plasticity at neocortical PN synapses. Both, long-term potentiation (LTP) and depression (LTD) have a presynaptic locus at these synapses and likely involve changes in pN, but the final steps that alter release are not well understood (Feldman, 2009; Castillo, 2012; Feldman, 2012). A recent study on L5PN synapses in S1 showed an increase in the number of primed vesicles as a major mechanism of LTP (Weichard et al., 2023). PN synapses in S1 are already tightly coupled and therefore a further decrease in the coupling distance is an unlikely mechanism of LTP. However, this does not exclude that in S1 LTD induction breaks tight coupling and vice versa that LTP in PFC involves tightening of coupling. In addition, in PFC we found an increased contribution of NMDAR on the postsynaptic site. Thus, these synapses have increased potential for plasticity on the pre- and the postsynaptic site. During induction of long-term plasticity, the postsynaptic site signals back to the presynaptic site via retrograde messengers (Regehr et al., 2009) and loose coupling will provide the substrate for enhanced regulatory capacity (Vyleta and Jonas, 2014).

In general, loose coupling provides more flexibility to regulate [Ca2+]i at the release sensor (Vyleta and Jonas, 2014). Prefrontal networks carry out higher-order computations that transform sensory information and memory to perform flexible output such as delayed response, inference and planning (Narayanan et al., 2025). Our findings suggest that loose coupling provides a synaptic foundation for such flexible neocortical network functions.

Methods

Slice preparation

C57BL/6J mice at P21-26 of either sex were used in this study. All experiments were performed in accordance with the guidelines for the welfare of experimental animals issued by the European Communities Council Directive (2010/63/EU) and with the German Protection of Animals Act (Tierschutzgesetz). Mice were bred in the animal facility of the Medical Faculty of Leipzig University and were housed in individually ventilated cages in a specific pathogen free environment and in a 12h/12h light dark cycle with access to food and water ad libitum. Experiments were approved by the animal welfare office of the University Medical Center, Leipzig, as well as the local governmental authorities (Landesdirektion Leipzig, registration numbers T10/20, T05/21-MEZ).

C57BL/6J mice at P21-26 of either sex were decapitated under deep Isoflurane (Curamed) inhalation anesthesia. The brain was excised rapidly and placed in cooled (0-4°C) artificial cerebrospinal fluid (ACSF) containing (in mM): 125 NaCl, 2.5 KCl, 1.25 NaH2PO4, 26 NaHCO3, 1 MgCl2, 2 CaCl2, and 20 glucose, equilibrated with 95% O2 and 5% CO2 (pH 7.3-7.4). Coronar neocortical slices (150-250 μm thick) were cut from the PFC or S1 region (Figure 1A) with a vibratome (HM 650 V, Microm). For some paired recordings between L5PNs and for calcium imaging experiments parasagittal slices (200 µm) were prepared from PFC region. Slices were incubated for 30 min at 35°C and subsequently stored at room temperature. For experiments, slices were transferred to a recording chamber and continuously perfused with ACSF (2-3 ml per min, supplemented with 10 μM (-)-bicuculline methiodide (Tocris) at 31-33°C. Unless stated otherwise, chemicals were from Sigma-Aldrich.

Electrophysiological recordings

Patch-clamp recordings from L5PNs were established as described in detail previously (Bornschein et al., 2019a; Bornschein et al., 2025). Patch pipettes were prepared from borosilicate glass (Hilgenberg) with a PC-10 puller (Narishige) and had final resistances of 6-8 MΩ when filled with the following standard pipette solution (in mM): 150 K-gluconate, 4 NaCl, 3 MgCl2, 3 Na2ATP, 0.3 NaGTP, 0.05 EGTA, 10 KHEPES, dissolved in purified water. The pH was adjusted to 7.3 with KOH. Recordings were performed under optical control (BX51WI, Olympus), using an EPC10/2 amplifier and Patchmaster software (version v2x90.2, HEKA). EPSCs were recorded in the whole-cell configuration at a holding potential (Vhold) of -90 mV (online corrected for a liquid junction potential of 16 mV), filtered at 5 kHz and sampled at 10 kHz. Series resistance (Rs) was continuously compensated to a fixed value between 10 and 15 MΩ. The average uncompensated Rs was 18±1 MΩ in L5PN-L5PN paired recordings (n=41 cells, mean ± SEM) and 23±1 MΩ in L2/3-L5PN recordings (n=133 cells). Holding current (Ihold) was also monitored continuously and was <180 pA (-134±16 pA in L5PN-L5PNs; -168±11 in L2/3-L5PNs). Presynaptic L5PNs were stimulated in on-cell configuration (200-500 mV, 1-2 ms).

For chelator wash-in experiments, presynaptic neurons were stimulated in the whole-cell configuration after re-patching with a pipette solution supplemented with 10 mM EGTA (K-gluconate concentration was reduced to 135 mM to adjust osmolarity). EPSCs were evoked every 10 s for at least 30 min (Figure 2). Averaged EPSC amplitudes were calculated before (≥10 min baseline) and 20-30 min after repatching (test period). For recordings from L2/3-L5PN connections ACSF-filled patch pipettes were placed close to the somata of PNs in L2/3 and extracellular stimulation was performed using an ISO-Stim 01 DPI (NPI electronics). Minimal stimulation (0.5 to 5 V) was used and yielded EPSC amplitudes of ∼40 pA (47±3 pA, n=131 pairs), making it likely that we investigated transmission mainly between single or few L2/3PNs and L5PNs in these experiments. Following 10 min of stable recordings in normal ACSF (baseline), the bath solution was exchanged for ACSF containing either 0.1% DMSO and 0.01% Pluronic (control solution), 10 µM EGTA-AM or 10 µM BAPTA-AM in DMSO/Pluronic for 30 min (incubation period) and thereafter replaced by normal ACSF again for another 10 min (test period). Averaged EPSC amplitudes were calculated during baseline and test period. The time courses of chelator effects on EPSC amplitudes were analyzed by binning EPSC amplitudes within 2 min intervals of recording time to average amplitudes. Chelator effects were expressed as averaged EPSC amplitudes during test period normalized to the corresponding baseline values.

VGCCs were inhibited by bath-application of ACSF containing a subsaturating concentration of 5 µM CdCl2 (Figure 4J). NMDA receptors were blocked by the selective antagonist 2-Amino-5-phosphonovaleriansäure (APV, 50 μM; SI Appendix, Figure S2). For assessing the effects of the individual blockers, EPSCs were recorded in ACSF for at least 10 min (10 s intervals), subsequently the blocker was perfused and the blocker effects were quantified after a stable block had been established (at least 5 min test period).

Quantification of quantal synaptic parameters

Quantal synaptic parameters were estimated from parabolic fits to mean-variance (MV) plots of EPSC amplitudes recorded at different [Ca2+]e (1, 2, and 5 mM, ≥30 repetitions per concentration; [Mg2+]e was correspondingly adjusted to 2, 1, 0 mM, respectively), assuming binominal release statistics (Clements and Silver, 2000; Scheuss et al., 2002; Silver, 2003). The recordings always started in 2 mM [Ca2+]e. The variance of EPSCs (σ2) was calculated according to ref. (Scheuss et al., 2002), plotted against the averaged amplitudes (I), and fitted by a parabola of the form: (1) σ2=Iq−I2N(1+CVII2)+qICV12 where q is the quantal size, N a binominal parameter, and CVI and CVII the coefficients of intrasite and intersite quantal variability, assumed to be 0.3 (Clements and Silver, 2000). The fits were constrained to pass through zero. The variance of the variance was calculated according to ref. (Meyer et al., 2001). Stationarity of EPSC amplitudes was established when, after full exchange of the bathing solutions (≥5 min), EPSC amplitudes no longer showed a tendency to increase or decline. For MV analysis in L2/3- L5PN connections, the bath solution was supplemented with the competitive AMPA receptor antagonist kynurenic acid (Kyn, 0.25 mM), which relieves their desensitization and saturation (Neher and Sakaba, 2001) and with 50 µM APV to prevent NMDAR activation (Figure 3). Due to the small size of EPSCs in L5PN-L5PN connections no blocker was used in these recordings. The size of qEPSCs was quantified from evoked miniature EPSCs recorded in low [Ca2+]e (1 mM) and high [Mg2+]e (2 mM).

Ca2+ imaging

Action potential-evoked (current injection of 1-4 nA for 1-2 ms) fluorescence changes were recorded at boutons located on axon collaterals of L5PNs as described previously (Bornschein et al., 2019a; Bornschein et al., 2025). L5PNs were filled with EGTA-free, Fluo-5F (200 µM, Invitrogen) and Alexa-594 (50 µM, Molecular Probes) containing pipette solution via somatic whole-cell patch-pipettes. Cells were dialyzed for at least 20 min to yield sufficient equilibration with the dyes. Volume averaged fluorescence signals were recorded in point mode at 500 kHz temporal resolution and subsequently binned to 500 Hz, using a custom-build two-photon microscope based on a Fluoview-300 scanner (Olympus), a 60x/0.9

N.A. objective, a mode-locked Ti:sapphire laser (Tsunami, Newport-Spectra Physics, set to a center wavelength of 810 nm), and a Pockels cell (350-80 KD*P, Conoptics). The fluorescence background was determined in a subsequent point mode recording performed near the boutons under investigation. Typically, at least 5 boutons per L5PN were recorded and averages per cell were calculated. The fluorescence signals were filtered (HC647/75, Semrock HC525/50, 720-SP, AHF), detected by two external PMT modules (H7422-40, Hamamatsu; PMT-02M/PMM-03, NPI electronics) monitoring red and green epifluorescence, respectively at fixed PMT voltages, and digitized with the Fluoview system. The Ca2+-dependent green fluorescence was normalized to the Ca2+-insensitive red fluorescence and expressed as background-corrected ΔG/R (Sabatini et al., 2002). ΔG/R signals were converted to changes in [Ca2+]i based on an in vitro quantification of the KD of Fluo-5F (439 nM) in our pipette solution, adjusted with CaEGTA and K2EGTA (100 mM stock solutions, 10 mM HEPES added) to contain different free Ca2+ concentrations that were calculated with the MaxChelator (https://somapp.ucdmc.ucdavis.edu/pharmacology/bers/maxchelator/CaMgATPEGTA-NIST.htm). Rmin and Rmax values were measured after each successful recording in sealed patch pipettes that were placed in the slices near the recorded cells. The pipettes contained the EGTA-free, Fluo-5F- and Alexa-594-containing pipette solution with either 0 mM CaCl2 and 10 mM K2EGTA (zero Ca2+ solution) or 20 mM CaCl2 (high Ca2+ solution).

Modeling

The kinetic gating scheme of the VGCCs and the reaction schemes for Ca2+ binding to buffers and transmitter release were converted to ordinary differential equations (ODEs). The system of ODEs was numerically solved using “NDSolve” or “NDSolveValue” of Mathematica 14 (Wolfram) as described previously (Schmidt et al., 2013; Bornschein et al., 2019a; Bornschein et al., 2025). Spatial resolution was achieved by placing the ODEs in concentric hemi-shells (1 or 2 nm thickness) that were coupled via diffusion.

The presynaptic action potential was modeled with a half-duration of 0.2 ms, τrise of 0.1 and τdecay of 0.2 ms (Borst et al., 1995). Voltage dependent activation of VGCCs was simulated using the gating scheme for P/Q-type channels, a single channel conductance of 2.5 pS, and a Ca2+ equilibrium potential of 130 mV (Li et al., 2007). The model included ATP as mobile buffer and immobile endogenous buffers with a buffer capacity (κE) of 25 (Tran and Stricker, 2018). Model parameters are given in Table S1 (SI Appendix).

The model for each channel consisted of five closed (C) and one open (O) state. The transition between the first five steps was voltage dependent and the last step voltage independent: (2) C0⇄α1(V)β1(V)C1⇄α2(V)β2(V)C2⇄α3(V)β3(V)C3⇄α4(V)β4(V)C4⇄ab0 a and b are the rate constants for transitions between C4 and O, and αi(V) and βi(V) are the voltage dependent forward and backward transition rates that are given by: (3a) αi(V)=αi,0exp⁡(V/ki)i=1−4 (3b) βi(V)=βi,0exp⁡(−V/kj)i=1−4 where αi,0 and βi,0 are the forward and backward rate constants at 0 mV and ki a slope factor.

Ca2+ binding to all buffers (B) was simulated by second order kinetics with forward and backward binding rate constants kon and koff, respectively: (4a) d[Ca]dt=−kon[Ca][B]+koff[CaB] (4b) d[B]dt=−d[CaB]dt Diffusion of all species (X) was simulated according to (5) d[X]dt=DxAn,n±1Vnr([X]n±1−[X]n) where DX is the diffusion coefficient, A the surface, V the volume and r the radius of a shell (n). Ca2+ was cleared by a linear, surface-based extrusion mechanism driven by the difference between [Ca2+]i and the resting [Ca2+]i (30 nM according to the imaging data).

For fitting the model to the average experimental Ca2+ transient, the number of VGCCs was adjusted, yielding 33 VGCCs (Figure 4F). The other free parameter of the simulation was the extrusion rate, which however affected only the later phase of the decay of the transient. During fitting, the model included the Ca2+ indicator dye. Fitting was first performed in a single compartment model (Helmchen and Tank, 2005). Subsequently, fitting was performed in the spatially resolved model with the numbers of VGCCs remaining unaltered and [Ca2+]i as reported by the dye was calculated from the concentration of the Ca2+- dye complex assuming equilibrium conditions (Schmidt et al., 2003). Since L5PNs are characterized by an immobile and small κE with unknown binding kinetics, a slow (EBs) and a fast immobile endogenous buffer (EBf) were assumed (Helmchen et al., 1996; Ohana and Sakmann, 1998; Tran and Stricker, 2018). The fractions of EBs and EBf were adjusted to optimize the fit to the measured Ca2+ transient and to sum up to κE of 25. Following fitting, the indicator dye was removed from the model and the recently developed model for release from L5PN terminals was included (Bornschein et al., 2025). Either the “Syt1 priming model” (Equation 6; Figure 4) or the “Syt1 model” (Equation 7; SI Appendix, Figure S4) were placed at varying distances from the Ca2+ sources. In the priming model (Equation 6), a single Ca2+-dependent priming step with Michaelis-Menten kinetics (KM) and a priming rate kprim is assumed. V is the vesicular release sensor of a primed vesicle (Equation 6, 7) that binds a maximum of 5 Ca2+ ions. kon and koff are the forward and backward Ca2+ binding rate constants, βi (i=0-4) is a cooperativity factor, and γ the fusion rate. pN was obtained by integrating the release rates over time.

A ring like arrangement of the VGCCs around a vesicle was assumed (Figure 4I; ring 1 model in ref. (Bornschein et al., 2019a)). The number of VGCCs driving release was continuously reduced from the maximal value of 33 (Figure 4F) and the coupling distances at which the simulated pN values matched the experimentally determined pN were obtained. Subsequently, EGTA was included and the consistency between the experimentally quantified EGTA effect and the simulated EGTA effect at a given coupling distance was determined (Figure 4G). The best match between experiments and simulation was obtained for 9 VGCCs at a coupling distance of 51 nm (Figure 4H).

Quantification and statistical analysis

Data from L5PN-L5PN pairs in PFC were compared to previously published data from these pairs in S1 (Bornschein et al., 2019b; Bornschein et al., 2019a). Electrophysiological data and Ca2+ imaging data were analyzed using custom written routines in Igor Pro (version 6.32 and 8.03, Wavemetrics). Data are shown as median and interquartile range (IQR) or as mean ± standard error (SEM).

Summarized data are either shown in boxplots as median ± IQR and mean values included as dashed lines in addition or in bar graphs as mean ± SEM. Normality was tested using the Shapiro-Wilk Test. Normally distributed data were compared with the t-test (two groups) or an ANOVA (more than two groups). Non-normally distributed or small samples of data were compared with the Mann-Whitney-U rank sum test (MWU; two groups) or an ANOVA on ranks (more than two groups). To compare pre- and post-treatment data the paired t-test or the Wilcoxon signed rank test (WSR) was used, depending on the distribution of the data. All statistical tests were two-tailed. P values are indicated as *P<0.05, **P< 0.01 and ***P<0.001. The number of experiments n represents the number of cell pairs or cells and was chosen sufficiently high to permit reliable statistical analysis. All of the statistical details of experiments can be found in the figures and corresponding figure legends. Statistics were performed with Sigma Plot 11.0 (Dundas Software) and Mathematica 14.

eLife Assessment

By using a combination of patch clamp recordings, calcium imaging and computer modeling, the authors analyze the spatial distribution of voltage gated calcium channels at glutamatergic synapses formed between layer 5 pyramidal neurons (L5PNs) and between layer 2/3 and L5PNs in the prefrontal cortex (PFC) and primary somatosensory cortex (S1); they conclude that the calcium channel-vesicle coupling is looser in the PFC compared to S1, although additional experiments are needed to determine how the distinct functional characteristics of these synapses in different brain regions might affect data interpretation. Overall, these findings are important because they have implications for shaping synaptic plasticity and neural circuit function across brain regions. They are solid because they are based on the use of a multi-pronged approach, although the presentation would benefit from stronger integration of the current findings with the existing literature and a more explicit discussion of potential limitations and confounding factors for data interpretation.

Reviewer #1 (Public review):

Summary:

This study asks whether synapses formed by the same broad neuronal class (excitatory pyramidal neurons, PN) adapt their presynaptic organization in a cortex-specific manner, comparing the prefrontal cortex (PFC) with the primary somatosensory cortex (S1). The authors combine sophisticated electrophysiology (paired recordings and extracellular minimal stimulation), pharmacological perturbations of presynaptic Ca²⁺-secretion coupling, bouton Ca²⁺ imaging, and mechanistic modeling. Across two prominent excitatory connections (Layer 5 (L5) PN-L5PN and L2/3-L5PN), they provide convergent evidence that mature PFC synapses operate with looser Ca²⁺ channel-release sensor coupling than their S1 counterparts.

Overall, the study provides an appealing mechanistic link between synaptic nano/micro-architecture and cortical-area specialization. The idea that PFC synapses retain a more "plasticity-favoring" presynaptic state, while the primary sensory cortex emphasizes reliability and timing precision, is potentially impactful for how we think about circuit computation and plasticity across cortical hierarchies.

Strengths:

A major strength is the multi-pronged experimental strategy. The paper first establishes robust, area-dependent differences in synaptic efficacy, reliability, timing, and short-term plasticity (facilitation prevailing in PFC versus depression in S1), using both paired recordings and minimal extracellular stimulation paradigms. The coupling interpretation is then directly supported by differential sensitivity to EGTA (and appropriate positive-control effects of fast chelators). Finally, volume-averaged calcium signals are reported to be similar across areas, arguing against trivial explanations based on gross differences in calcium influx, and the modeling provides a quantitative framework for interpreting the observed chelator effects.

Weaknesses:

Limitations are minor and concern interpretation/clarity rather than core results. Some key inferences rely on indirect readouts (chelator sensitivity, fluctuation analysis-derived parameters, bouton-averaged calcium signals), each of which carries assumptions and potential confounds that should be discussed more explicitly. In particular, the repatching paradigm for the paired-recording EGTA experiment, though very impressive, and the limited number of extracellular calcium conditions used for fluctuation analysis (three concentrations), can influence quantitative estimates and the confidence intervals around them.

Reviewer #2 (Public review):

Schwarze et al. investigated whether synaptic efficacy is brain-region specific. To this end, they compared synaptic connections established by layer 5 (L5) neocortical pyramidal cells and between L5 and L2/3 pyramidal cells. In order to identify the mechanism of this brain region specificity, the authors employed several experimental approaches, including paired electrophysiological recordings, extracellular stimulation, low- and high-affinity intracellular calcium chelators (EGTA and BAPTA), multiple probability fluctuation analysis (MPFA), and intracellular measurements of calcium transients as well as computational modelling. The findings of the present study indicate that synaptic connections in the primary somatosensory cortex (S1) are significantly stronger and more reliable than those in the prefrontal cortex (PFC).

The study is timely, and the topic is of significant interest to the neuroscience community. Despite the extensive research that has been carried out on the neuroanatomy and receptor distribution of different brain regions, comparatively little attention has been paid to differences in synaptic physiology. The authors' approach is characterised by its elegance and comprehensive nature, and the conclusions drawn are compelling. Nevertheless, there are a number of unresolved issues.

Major points:

(1) The authors state that data from the S1 cortex were obtained in a previous study. In the context of an explicitly comparative study (PFC vs. S1cortex), it would have been advantageous for the authors to perform a subset of experiments in which both cortices were obtained from a single animal. This is a feasible undertaking, given the spatial separation of the PFC and S1 cortex.

(2) Figure 1A is somewhat misleading because it could suggest that the authors have performed dual recordings in identified PFC pyramidal cells.

(3) PFC and S1 cortex in rodents differ markedly in their morphological organisation. For example, in all sensory cortices, layer 4 is very pronounced; however, in the PFC of rodent,s no clear layer 4 can be found. On the other hand, PFC shows a clear separation of layers 2 and 3, which is not visible inthe S1 cortex. Furthermore, PFC pyramidal cells in layers 2, 3, and 5 exhibit significant heterogeneity, diverging considerably from those found in layers 5a and 5b of S1 cortex. Thus, there is no clear correlation between L5 pyramidal cells in the PFC and the S1 cortex. In order to achieve a meaningful comparison of the data obtained in PFC and S1 cortex, it is necessary for the authors to determine whether the record is from similar pyramidal cell populations.

(3) In addition, PFC pyramidal cells in layer 2, 3 and 5 are highly heterogeneous and differ markedly from those in layer 5a and 5b of S1 cortex. To achieve a meaningful comparison of the data obtained in the PFC and the S1 cortex, the authors need to determine whether the record from similar pyramidal cell populations.

(4) For the S1 cortex, in rats it has been found that L5 synaptic connection between pairs of L5a pyramidal cells and pairs of L5b pyramidal cells differ markedly with respect to mean EPSP amplitude, latency and coefficient of variation (cv, a surrogate measure for the synaptic release probability) (cf. Markram et al., 1997; Frick et al., 2008). It is therefore likely that PFC and S1 pre- and postsynaptic pyramidal cells are not only morphologically and electrophysiological distinct but also with respect to their synaptic properties. At least, the authors need to discuss these confounding issues and preferentially address them experimentally. For example, it would be helpful to demonstrate that paired recordings were made from the same pyramidal cell types, perhaps by documenting their morphology and/or firing patterns. In addition, they should discuss the marked difference in EPSP amplitude and putative release probability between their data and the earlier studies.

(5) In order to perform multiple probability fluctuation analysis (MPFA), a parabolic fit with a mere three points is inadequate, particularly because 2 mM and 5 mM Ca2+ are close to the peak of the variance-to-mean parabola, and only 1 mM Ca2+ is on its initial linear part. A more meaningful result would have been obtained with an additional Ca2+ concentration between 1.0 and 2.0 mM, as these are closer to the physiological range. In this context, the authors should have quoted the more recent and more detailed paper by the Silver group (Saviane and Silver, 2006; Lanore and Silver, 2016) and not just the Clements and Silver review paper.

(6) Methods: The authors should clarify whether their paired recordings from L5 pyramidal cells involved whole-cell recordings from both pre- and postsynaptic neurons. From Figure 1B, it appears as if the presynaptic neurons were not recorded in whole cell mode but rather stimulated in cell-attached mode. This is also reflected in the artefact visible in the current trace recorded in the postsynaptic neuron. The authors should explicitly state their methodological approach and mention how reliable the timing of the presynaptic action potential was under these circumstances. The same holds true for the extracellular stimulation protocol. A significantly more detailed description of the experimental protocol is necessary here.

(7) Methods: The authors use Student's t-test for data comparison. The authors should verify that the data distribution was indeed normal, e.g. by using a Shapiro-Wilk test. If this is not the case, non-parametric tests should be used.

Reviewer #3 (Public review):

Summary:

In this manuscript, Max Schwarze and colleagues examined the coupling distance between presynaptic Ca²⁺ channels and the vesicular release sensor at neocortical synapses in mice. They propose that Ca²⁺ channel-release sensor coupling differs across cortical areas, with relatively loose (microdomain) coupling in prefrontal cortex (PFC) and tighter (nanodomain) coupling in primary somatosensory cortex (S1) for comparable pyramidal-neuron synapse types. To test this, they combine paired recordings and minimal stimulation with chelator manipulations (EGTA/BAPTA), mean-variance/MPFA-style analyses, presynaptic Ca²⁺ imaging, and computational modeling. They conclude that presynaptic coupling organization is area-specific in the mature cortex and contributes to regional differences in synaptic timing, reliability, and short-term plasticity.

Strengths:

This study tackles an important question and is strengthened by a cohesive body of evidence assembled from multiple complementary approaches. A major asset is the inclusion of high-value datasets, particularly the paired recordings between L5 pyramidal neurons and the systematic assessment of EGTA sensitivity, which provide a solid functional foundation for the authors' central claims. The work is further distinguished by its genuinely multimodal design: combining electrophysiology with presynaptic calcium imaging (and integrating these observations with quantitative analyses and modeling) offers a more mechanistic view of neurotransmitter release than any single method could provide. Overall, the direct, within-framework comparison of presynaptic release-control mechanisms across cortical areas for comparable synapse types is compelling and gives the conclusions a level of robustness and interpretability that is often difficult to achieve in studies of cortical synaptic diversity.

Weaknesses:

Several aspects would benefit from clearer explanation, stronger integration with the existing literature, and a more explicit discussion of limitations and potential confounds. Without these additions, some conclusions remain speculative. Throughout the manuscript, the authors also often imply that different measurements reflect the same underlying synapse population. This is unlikely to be strictly true across all experiments and makes it difficult to integrate results from the various approaches into a single, unified set of functional synaptic properties. In addition, some statements-particularly those linking coupling mode to "higher-order neocortical functions"-appear broader than what is directly supported by the experiments and should be tempered or more precisely scoped.

Below, I list several topics that could help better frame the main findings of the present study and clarify how it relates to previously published work.

(1) The authors use EGTA sensitivity of EPSCs (together with additional metrics) to argue that S1 and PFC synapses differ in Ca²⁺ channel-release sensor coupling. While this is a plausible interpretation, EGTA effects are not uniquely determined by coupling distance and can also reflect differences in Ca²⁺ entry kinetics, action potential waveform, endogenous buffering/extrusion, or release-sensor/vesicle state. The authors use a constrained modeling approach, but the rationale for the different constraint sets is not fully clear from the current description. It would be helpful to expand and clarify the Methods section to explain how these constraints were defined, justified, and applied (and how alternative constraint choices would affect the results). In this context, the Abstract's broader claim that the study "reveals microdomain coupling as a presynaptic structure-function correlate of higher-order neocortical functions" appears overstated. Given the well-known diversity of cortical synapses even within a single region (e.g., synapses onto different interneuron subclasses or different PN cell types, extracortical sources like thalamus), the authors should clarify the intended scope: is the conclusion meant to apply broadly across synapse classes in S1 and PFC, or only to the specific connection type(s) examined here?

(2) The chelator logic is sound in principle, but the Discussion should more explicitly acknowledge standard caveats and alternative explanations. The authors partly address this by including presynaptic Ca²⁺ imaging and modeling, yet it would help to explain more clearly how the combination of (i) chelator sensitivity, (ii) presynaptic Ca²⁺ signals, and (iii) model constraints rules out-or substantially reduces the likelihood of-changes in AP waveform, Ca²⁺ influx kinetics, buffering/extrusion, or sensor/vesicle state as the primary drivers. In addition, recent hypotheses emphasizing vesicle priming and/or release-site occupancy as contributors to apparent EGTA sensitivity should be discussed as a complementary or alternative interpretation.

(3) A substantial portion of the S1 comparison appears to rely on previously published datasets. This should be made unambiguous in the Results and Methods, and it would be helpful to summarize this clearly (e.g., in a table indicating which figures/analyses use new data versus reanalysis of published data). If this information is already present, it should be highlighted more prominently.

(4) The modeling is informative, but the choice of a specific VGCC-release-site geometry and channel arrangement is not sufficiently justified. The manuscript adopts a particular spatial configuration, yet the rationale for selecting this geometry, rather than other plausible architectures discussed in the literature, is not clearly explained, nor is it meaningfully revisited in the Discussion. The authors should justify why the same organization is assumed across two distinct cortical areas and, ideally, include (or at a minimum discuss) a sensitivity analysis showing how key inferences (e.g., coupling distance and channel number) depend on the assumed geometry.

(5) The calcium imaging data are valuable, but given the diversity of synapses within each cortical layer, it is not clear that imaged boutons can be confidently assigned to the specific connection types being interrogated electrophysiologically. A substantial fraction of boutons likely corresponds to different postsynaptic targets (including interneurons and distinct pyramidal-cell classes), and this heterogeneity could complicate interpretation. This limitation should be discussed explicitly

(6) In unitary connections, the authors assess EGTA effects alongside other functional parameters (strength, delay, short-term plasticity), which is a major strength. However, for L2/3 to L5 connections, it appears that EGTA sensitivity was tested primarily using extracellular stimulation. Given anatomical and circuit differences between PFC and S1, extracellular stimulation may recruit different synapse populations across regions, potentially confounding regional comparisons of EGTA sensitivity. This limitation should be acknowledged explicitly. While I am not requesting technically demanding L2/3↔L5 paired recordings in S1, the possibility that different synapse identities are being sampled should be treated as a meaningful source of uncertainty. The Discussion would also benefit from placing the magnitude of EGTA effects in the context of prior "loose coupling" literature, where comparatively large EGTA effects have been reported in some systems. In addition, the reported difference between adult PFC EGTA effects and S1 inhibition appears small (on the order of <10%) and should be interpreted cautiously, especially given that PFC and S1 mature on different timelines and P21-P26 is unlikely to reflect a mature PFC circuit state. The adult cohort (P90-P100) is therefore important, but the age mismatch complicates PFC-S1 comparisons; ideally, S1 should be assessed at matched ages, or this limitation should be discussed explicitly. Finally, for statistical robustness, in panel D of Figure 2, were the comparisons corrected for multiple testing to control Type I error?

(7) Alterations in initial release probability are often associated with changes in short-term plasticity. In the present manuscript, the authors report similar initial release probability at PFC and S1 synapses, yet observe differences in short-term plasticity profiles. The mechanistic basis for this apparent dissociation is not addressed and should be discussed explicitly, including potential explanations.

(8) There are multiple instances where the text appears to cite non-existent or misnumbered figure panels (e.g., references to "Figure 4G-I / 4J" when the relevant material appears elsewhere). These should be corrected throughout, as they currently reduce readability and confidence.

(9) The Methods describe P21-P26 animals, whereas the Results include older cohorts (e.g., P90-P100) and additional regions (e.g., mPFC). The Methods should be updated so that all cohorts and regions analyzed in the Results are fully described.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

This study asks whether synapses formed by the same broad neuronal class (excitatory pyramidal neurons, PN) adapt their presynaptic organization in a cortex-specific manner, comparing the prefrontal cortex (PFC) with the primary somatosensory cortex (S1). The authors combine sophisticated electrophysiology (paired recordings and extracellular minimal stimulation), pharmacological perturbations of presynaptic Ca2+-secretion coupling, bouton Ca2+ imaging, and mechanistic modeling. Across two prominent excitatory connections (Layer 5 (L5) PN-L5PN and L2/3-L5PN), they provide convergent evidence that mature PFC synapses operate with looser Ca2+ channel-release sensor coupling than their S1 counterparts.

Overall, the study provides an appealing mechanistic link between synaptic nano/micro-architecture and cortical-area specialization. The idea that PFC synapses retain a more "plasticity-favoring" presynaptic state, while the primary sensory cortex emphasizes reliability and timing precision, is potentially impactful for how we think about circuit computation and plasticity across cortical hierarchies.

Strengths:

A major strength is the multi-pronged experimental strategy. The paper first establishes robust, area-dependent differences in synaptic efficacy, reliability, timing, and short-term plasticity (facilitation prevailing in PFC versus depression in S1), using both paired recordings and minimal extracellular stimulation paradigms. The coupling interpretation is then directly supported by differential sensitivity to EGTA (and appropriate positive-control effects of fast chelators). Finally, volume-averaged calcium signals are reported to be similar across areas, arguing against trivial explanations based on gross differences in calcium influx, and the modeling provides a quantitative framework for interpreting the observed chelator effects.

Weaknesses:

Limitations are minor and concern interpretation/clarity rather than core results. Some key inferences rely on indirect readouts (chelator sensitivity, fluctuation analysis-derived parameters, bouton-averaged calcium signals), each of which carries assumptions and potential confounds that should be discussed more explicitly. In particular, the repatching paradigm for the paired-recording EGTA experiment, though very impressive, and the limited number of extracellular calcium conditions used for fluctuation analysis (three concentrations), can influence quantitative estimates and the confidence intervals around them.

We would like to thank the reviewer for his/her overall positive assessment of our manuscript and the constructive advice, which helped us to improve our manuscript. We discussed the limitations, assumptions and potential confounding factors in more detail. We addressed them pointwise in the recommendations for the authors.

Reviewer #2 (Public review):

Schwarze et al. investigated whether synaptic efficacy is brain-region specific. To this end, they compared synaptic connections established by layer 5 (L5) neocortical pyramidal cells and between L5 and L2/3 pyramidal cells. In order to identify the mechanism of this brain region specificity, the authors employed several experimental approaches, including paired electrophysiological recordings, extracellular stimulation, low- and high-affinity intracellular calcium chelators (EGTA and BAPTA), multiple probability fluctuation analysis (MPFA), and intracellular measurements of calcium transients as well as computational modelling. The findings of the present study indicate that synaptic connections in the primary somatosensory cortex (S1) are significantly stronger and more reliable than those in the prefrontal cortex (PFC).

The study is timely, and the topic is of significant interest to the neuroscience community. Despite the extensive research that has been carried out on the neuroanatomy and receptor distribution of different brain regions, comparatively little attention has been paid to differences in synaptic physiology. The authors' approach is characterised by its elegance and comprehensive nature, and the conclusions drawn are compelling. Nevertheless, there are a number of unresolved issues.

First, we would like to thank the reviewer for his/her detailed survey of our work, which was very helpful in improving our manuscript. We are happy about the overall positive evaluation and the constructive comments. To fully clarify all points, we performed new experiments and analyses, in particular we determined EGTA sensitivity in PFC and S1 from the same animal and we performed MPFA with an additional extracellular Ca2+ concentration. We extended the discussion on the examined cell types. Overall, we carefully revised the manuscript to address all points. Please see below our point-wise response.

Major points:

(1) The authors state that data from the S1 cortex were obtained in a previous study. In the context of an explicitly comparative study (PFC vs. S1cortex), it would have been advantageous for the authors to perform a subset of experiments in which both cortices were obtained from a single animal. This is a feasible undertaking, given the spatial separation of the PFC and S1 cortex.

This is only true for the paired recordings from L5PN-L5PN connections in S1, which were obtained in a previous study and partially reanalyzed. All recordings from L2/3-L5PN connections in S1 and PFC as well as the paired recordings on L5PN-L5PN synapses in PFC were obtained in the present study. To make this clearer, we have added Table 1. This lists which data and associated figures are from this study and which are from previous studies (Bornschein et al., Cell Rep. 2019; Bornschein et al., Front. Syn. Neurosci. 2019).

Our experiments are lengthy and therefore it is challenging to achieve two successful recordings within the lifetime of acute brain slices. For this reason, the previous version of the manuscript did not include recordings from PFC and S1 of the same animal. We have now measured EGTA effects in L2/3-L5PNs from S1 and PFC of the same animal. Two new recordings were added to the EGTA-AM plots in Figure 2C-F and in the results section. Example recordings are shown in Figure S3A and B, as is the comparison of EGTA-AM effects in L2/3-L5PNs from PFC and S1 in Figure S3C, with data points from the same animal marked.

“On the other hand, EGTA significantly reduced EPSC amplitudes only in PFC (0.54, 0.47-0.67; 55% of control) but not in S1 (0.85, 0.81-1.04; 100% of control). For a better comparison of EGTA effects some recordings were performed in PFC and S1 derived from the same animal to rule out interindividual effects (see example recording in Figure S3A-C).”

(2) Figure 1A is somewhat misleading because it could suggest that the authors have performed dual recordings in identified PFC pyramidal cells.

We thank the reviewier for this helpful note. We added “L2/3 or L5” to the stimulation panel of Figure 1A to illustrate that we stimulated either extracellularly in L2/3 or L5PNs directly via the patch pipette.

(3) PFC and S1 cortex in rodents differ markedly in their morphological organisation. For example, in all sensory cortices, layer 4 is very pronounced; however, in the PFC of rodent,s no clear layer 4 can be found. On the other hand, PFC shows a clear separation of layers 2 and 3, which is not visible inthe S1 cortex. Furthermore, PFC pyramidal cells in layers 2, 3, and 5 exhibit significant heterogeneity, diverging considerably from those found in layers 5a and 5b of S1 cortex. Thus, there is no clear correlation between L5 pyramidal cells in the PFC and the S1 cortex. In order to achieve a meaningful comparison of the data obtained in PFC and S1 cortex, it is necessary for the authors to determine whether the record is from similar pyramidal cell populations.

(3) In addition, PFC pyramidal cells in layer 2, 3 and 5 are highly heterogeneous and differ markedly from those in layer 5a and 5b of S1 cortex. To achieve a meaningful comparison of the data obtained in the PFC and the S1 cortex, the authors need to determine whether the record from similar pyramidal cell populations.

We apologize for having not been precise about the specific location and type of pyramidal neurons in the original manuscript. Extracellular stimulation in PFC and S1 was always performed in layer 2, where the first large cell bodies, relative to the pia mater, are located within a cortical column. Therefore, we assume that the same cell populations were stimulated in both brain regions (van Aerde & Feldmeyer, Cereb. Cortex 2015; Oberlaender et al., Cereb. Cortex 2012; Lefort et al, Neuron 2009). To stick with the standard terminology, we refer to it in the manuscript as upper layer 2/3. We kept stimulation intensity as low as possible to ensure that only a few presynaptic cells within the target region were activated.

In S1, paired recordings were obtained from pyramidal neurons in layer 5A following the procedures and criteria described in detail in our previous work (Bornschein et al., Cell Rep. 2019; Bornschein et al., Front. Syn. Neurosci. 2019; Bornschein et al., Science 2025). Briefly, these criteria are as follows: close proximity to layer 4 and the barrels as well as the PPR of 0.78 (0.69-0.90), which is consistent with depression dominating in L5A (Frick et al., Cereb. Cortex 2008; Bornschein et al., Front. Syn. Neurosci. 2019) and different from L5B with PPR ≥ 1 (Lefort & Petersen, Cereb. Cortex 2017). Within layer 5A we did not attempt to further differentiate between pyramidal neuron types. For recordings in S1 with extracellular stimulation we focused on the same locations as for the paired recordings. We extended the corresponding section in Materials and Methods of the revised manuscript.

We agree that there is no clear layer 4 in PFC, making the distinction between layer 2/3 and layer 5 less clear. Layer 2/3 and layer 5 have approximately the same diameter (van Aerde & Feldmeyer, Cereb. Cortex 2015). Based on this, we performed recordings in upper layer 5 of PFC. We did neither morphologically nor electrophysiologically differentiate between pyramidal neuron cell types. It should be noted that within a cortical area (S1 or PFC), we did not find a difference between glutamatergic synapses from L2/3 onto L5PNs and L5PN-to-L5PN synapses, neither with regard to the EGTA-sensitivity of release nor with regard to the release probability. In particular, we found homogeneous results and similar variability in both, the examined connections in the PFC and in S1, with no discernible clustering in the data that would suggest stimulation of different cell populations. These findings suggest that excitatory inputs to L5PNs exhibit similar properties (PPR, pN, CD) irrespective of whether they originate in L2/3 or in neighbouring PNs in L5A. However, we do see significant differences between synapses in the different cortical areas S1 and PFC. Thus, intra-area specific differences in morphology and spiking patterns among pyramidal neurons appear to not be reflected on the level of their synapses.

The Reviewer probably refers to such differences and heterogeneity in morphology and spiking patterns of pyramidal neurons. If he/she has more specific differences in mind, it would be helpful if references for the significant heterogeneity could be given.

Please also note that the type of experiments we perform with paired recordings and long-lasting patch-clamp measurements is not suitable for analyzing population differences among pyramidal neuron types.

We refer to the problem of pyramidal neuron heterogeneity in the revised manuscript in the discussion.

“Patch-clamp recordings from L5PNs located in the upper layer 5 (L5A in S1) were established according to the criteria described in detail in our previous work on this connection in S1 (Bornschein et al., 2019b; Bornschein et al., 2025). Presynaptic neurons were stimulated extracellularly in upper layer 2/3 (L2/3-L5PN connections) straight above the patched L5PN or in on-cell mode in L5A right next to the postsynaptic cell (L5PN-L5PN connections; Figure 1).“

“We did neither morphologically nor based on spiking patterns differentiate further between PN subtypes within a given layer. However, within a cortical area (S1 or PFC) we did not find a difference between glutamatergic synapses from L2/3 onto L5PNs and L5PN to L5PN synapses, neither with regard to the EGTA sensitivity of release nor with regard to pN. In particular, we found homogeneous results and similar variability in both, the examined connections in the PFC and in S1, with no discernible clustering in the data that would indicate stimulation of different cell populations. These findings suggest that excitatory inputs to L5PNs exhibit similar properties (PPR, pN, CD) irrespective of whether they originate in L2/3 or in neighboring PNs in L5A. However, we do see significant differences between synapses in the different cortical areas S1 and PFC. Thus, intra-area specific differences in morphology and spiking patterns among PNs appear to be not reflected on the level of their synapses.“

(4) For the S1 cortex, in rats it has been found that L5 synaptic connection between pairs of L5a pyramidal cells and pairs of L5b pyramidal cells differ markedly with respect to mean EPSP amplitude, latency and coefficient of variation (cv, a surrogate measure for the synaptic release probability) (cf. Markram et al., 1997; Frick et al., 2008). It is therefore likely that PFC and S1 pre- and postsynaptic pyramidal cells are not only morphologically and electrophysiological distinct but also with respect to their synaptic properties. At least, the authors need to discuss these confounding issues and preferentially address them experimentally. For example, it would be helpful to demonstrate that paired recordings were made from the same pyramidal cell types, perhaps by documenting their morphology and/or firing patterns. In addition, they should discuss the marked difference in EPSP amplitude and putative release probability between their data and the earlier studies.

We agree that Markram et al. (J. Physiol. 1997) and Frick et al. (Cereb. Cortex 2008) provided highly valuable insights into synaptic transmission between pyramidal neurons in S1. We referred to their work in detail in our previous work on developmental changes in the presynaptic organization of transmitter release in L5APN synapses in S1 (Bornschein et al., Cell Rep. 2019). Both studies were performed in young rats and EPSPs were measured, whereas we worked in mice and recorded EPSCs. This impedes a direct comparison of amplitudes.

Markram et al. (J. Physiol. 1997) recorded in 2-week-old rats from thick tufted PNs, corresponding to L5BPNs. Given the longer lifespan and slower development of rats compared to mice, this likely reflects a maturation state that corresponds better to our previous measurements in 8 to 10-day-old mice. Markram et al. found small failure rate (median 7%), which is similar to what we found in our previous study for young L5APN synapses (low failure rates and high pN; Bornschein et al., Cell Rep. 2019).

The study by Frick et al. (Cereb. Cortex 2008) is closer to our present study and to the mature age window in our previous study, although they also recorded from rats but from L5APN-L5APN pairs in almost 3-week-old animals in S1. Again, EPSPs rather than EPSCs were recorded, impeding a direct comparison of amplitudes.

Both studies concluded, based on the synaptic failure rate and CV analysis of EPSP amplitude, that the synapses they investigated operate with high release probability. This is fully in line with our findings. Of note, we found no significant difference between L2/3-L5PN and L5PN-L5PN synapses within a given area, indicating that varibality on the synaptic level between PNs of a given area is not pronounced.

In order to further substantiate this, we determined the relative variability in median EPSC amplitudes to test whether there is a higher variability of recorded cell types in PFC compared to S1. The relative MAD (median absolute deviation) of EPSC amplitudes was 0.46 in PFC and 0.50 in S1. The similarity in these values argues against higher cell-type variability in PFC compared to S1. We have discussed the results of these studies in relation to our own findings.

“Two other previous studies on L5APN (Frick et al., 2008) and L5BPN (Markram et al., 1997) connections concluded that these synapses operate with high release probability, which nicely agrees with our previous (Bornschein et al., 2019b) and current results. It is remarkable that we did not even detect any differences between the L2/3-L5PN and L5PN-L5PN synapses within a given cortical area. Overall these results from different studies (Markram et al., 1997; Reyes and Sakmann, 1999; Frick et al., 2008; Bornschein et al., 2019b; Bornschein et al., 2019a) may indicate that variability on the synaptic level between PNs of a given area is not pronounced. In order to further substantiate this, we determined the relative variability in median EPSC amplitudes to test whether there is a higher variability of recorded cell types in PFC compared to S1. The relative MAD (median absolute deviation) of EPSC amplitudes was 0.46 in PFC and 0.50 in S1. The similarity in these values argues against higher cell-type variability in PFC compared to S1.

(5) In order to perform multiple probability fluctuation analysis (MPFA), a parabolic fit with a mere three points is inadequate, particularly because 2 mM and 5 mM Ca2+ are close to the peak of the variance-to-mean parabola, and only 1 mM Ca2+ is on its initial linear part. A more meaningful result would have been obtained with an additional Ca2+ concentration between 1.0 and 2.0 mM, as these are closer to the physiological range. In this context, the authors should have quoted the more recent and more detailed paper by the Silver group (Saviane and Silver, 2006; Lanore and Silver, 2016) and not just the Clements and Silver review paper.

We used only three Ca2+ concentrations for MPFA as these resulted in a low (<0.5), a medium (~0.5) and a large (>0.5) pN condition, thereby clearly determining a parabola. Also Saviane and Silver (Nature 2006) performed MPFA with three extracellular Ca2+ concentrations (1, 2, and 8 mM). We have now cited this paper, as well as the more recent work by Lanore and Silver (Neuromethods 2016), in relation to the MPFA method. To verify the reliability of MPFA, the determined parameters were compared with values estimated based on the EPSC amplitudes (EPSC = N pN q; PFC, 6 pA; S1, 48 pA) and failure rates (F = (1-pN)^N; PFC, 0.25, S1, 0.0001). The estimated values did in fact match those of the MPFA (EPSCs in PFC: 8 pA, 5-15 pA, and S1: 29 pA, 18-53 pA; failure rates in PFC: 0.16, 0.08-0.28, and S1: 0, 0-0.03; see original manuscript.

To further support this, we have now conducted additional experiments using four Ca2+ concentrations. The results are consistent with those from the experiments using three concentrations. The additional Ca2+ concentration of 1.5 mM did not improve the parabolic fit, as it yielded pN values very close to those determined with 2 mM Ca2+. Therefore, the additional experiments are shown in Author response image 1. The novel pN data are included in the summary of pN values (now n=6) in the results section and in Figure 3F.

MPFA with four different extracellular Ca2+ concentrations.

(A) MPFA of EPSC amplitudes recorded at the indicated [Ca2+]e from L2/3-L5PNs in PFC. Top: Individual EPSCs (grey, average in black) recorded from L5PNs after extracellular stimulation in L2/3. Middle: Plot of EPSC amplitudes over time. Bottom: Corresponding mean-variance plot fitted with a parabola estimating the quantal parameters of release. pN is for 2 mM [Ca2+]e. Recordings were made in the presence of 10 µM Bicuculline, 0.25 mM Kynurenic acid and 50 µM 2-Amino-5-phosphonovaleriansäure. (B) As in (A), but for a L2/3-L5PN connection in S1. (C) Summary of determined pN values in PFC and S1 (dots represent individual experiments; P=0.485, Mann-Whitney U rank-sum test).

Additionally, we performed bootstrap analyses with 10,000 replicates which were generated with replacement from the original data sample. Distributions of bootstrap 25% trimmed means showed a clear separation for N and q between PFC and S1 but not for pN. The bootstrap results are shown in Author response image 2.

Bootstrap analysis.

(A-C) Distribution of bootstrap 25% trimmed means of the quantal parameters pN (A), N (B) and q (C) in PFC (orange), S1 (blue) and S1 with gDGG (light blue). 10,000 bootstrap replicates were generated with replacement from the original data sample obtained by MPFA in L5PN-L5PN connections (cf. Figure 3D). (D) Same as in (A) but for MPFA in L2/3-L5PN connections.

(6) Methods: The authors should clarify whether their paired recordings from L5 pyramidal cells involved whole-cell recordings from both pre- and postsynaptic neurons. From Figure 1B, it appears as if the presynaptic neurons were not recorded in whole cell mode but rather stimulated in cell-attached mode. This is also reflected in the artefact visible in the current trace recorded in the postsynaptic neuron. The authors should explicitly state their methodological approach and mention how reliable the timing of the presynaptic action potential was under these circumstances. The same holds true for the extracellular stimulation protocol. A significantly more detailed description of the experimental protocol is necessary here.

In the paired recordings, presynaptic cells were stimulated in the cell-attached mode. For presynaptic EGTA application the whole-cell configuration was established after re-patching to allow buffer perfusion of the presynaptic L5PN. This is described in the methods section of the original version of the manuscript. We extended this description as follows:

“In paired recordings, presynaptic L5PNs were stimulated in on-cell configuration (200-500 mV, 1-2 ms). In the chelator wash-in experiments, presynaptic neurons were repatched with a pipette solution supplemented with 10 mM EGTA (K-gluconate concentration was reduced to 135 mM to adjust osmolarity) and whole-cell configuration was established to allow EGTA perfusion of the presynaptic neuron.”

The amplitudes were determined by fitting a product of two exponential functions to the baseline-subtracted currents, which allows for independent adjustment of the time constants of the rising and falling phases and minimizes noise effects (cf. Bornschein et al., J. Physiol. 2013). Synaptic delays were determined from the onset of stimulation to the fitted EPSC onset. We added this more detailed explanation to the methods section. in the timing of the presynaptic action potential was similar in recordings from PFC and S1. This applies to the paired-recordings with on-cell stimulation of presynaptic neurons as well as to the extracellular stimulation experiments.

“Synaptic responses were determined by fitting a product of two exponential functions to the baseline-subtracted currents, which allows for independent adjustment of the time constants of the rising and falling phases and minimizes noise effects (Bornschein et al., 2013). Synaptic delays were determined from the onset of stimulation to the fitted onset of the EPSC. PPRs were calculated by dividing the second amplitude of two consecutive EPSCs by the first.”

(7) Methods: The authors use Student's t-test for data comparison. The authors should verify that the data distribution was indeed normal, e.g. by using a Shapiro-Wilk test. If this is not the case, non-parametric tests should be used.

We typically used non-parametric tests as stated in the figure legends of the corresponding figures. We have now explained the abbreviations for the Mann-Whitney U test (MWU) and the Wilcoxon signed-rank test (WSR) in the figure legends. A paired t-test was used only in Figure 5F after testing for normal distribution with the Shapiro-Wilk test. This is described in the methods section of the original version of the manuscript. Additionally, results of the Shapiro-Wilk test were now included in the figure legends.

“Normality was tested using the Shapiro-Wilk test. Normally distributed data were compared with the t-test (two groups) or a one-way ANOVA (more than two groups). Non-normally distributed or small samples of data were compared with the Mann-Whitney U rank-sum test (MWU; two groups) or a Kruskal-Wallis ANOVA on ranks (more than two groups). (…). To compare pre- and post-treatment data the paired t-test or the Wilcoxon signed-rank test (WSR) was used, depending on the distribution of the data.”

Reviewer #3 (Public review):

Summary:

In this manuscript, Max Schwarze and colleagues examined the coupling distance between presynaptic Ca2+ channels and the vesicular release sensor at neocortical synapses in mice. They propose that Ca2+ channel-release sensor coupling differs across cortical areas, with relatively loose (microdomain) coupling in prefrontal cortex (PFC) and tighter (nanodomain) coupling in primary somatosensory cortex (S1) for comparable pyramidal-neuron synapse types. To test this, they combine paired recordings and minimal stimulation with chelator manipulations (EGTA/BAPTA), mean-variance/MPFA-style analyses, presynaptic Ca2+ imaging, and computational modeling. They conclude that presynaptic coupling organization is area-specific in the mature cortex and contributes to regional differences in synaptic timing, reliability, and short-term plasticity.

Strengths:

This study tackles an important question and is strengthened by a cohesive body of evidence assembled from multiple complementary approaches. A major asset is the inclusion of high-value datasets, particularly the paired recordings between L5 pyramidal neurons and the systematic assessment of EGTA sensitivity, which provide a solid functional foundation for the authors' central claims. The work is further distinguished by its genuinely multimodal design: combining electrophysiology with presynaptic calcium imaging (and integrating these observations with quantitative analyses and modeling) offers a more mechanistic view of neurotransmitter release than any single method could provide. Overall, the direct, within-framework comparison of presynaptic release-control mechanisms across cortical areas for comparable synapse types is compelling and gives the conclusions a level of robustness and interpretability that is often difficult to achieve in studies of cortical synaptic diversity.

Weaknesses:

Several aspects would benefit from clearer explanation, stronger integration with the existing literature, and a more explicit discussion of limitations and potential confounds. Without these additions, some conclusions remain speculative. Throughout the manuscript, the authors also often imply that different measurements reflect the same underlying synapse population. This is unlikely to be strictly true across all experiments and makes it difficult to integrate results from the various approaches into a single, unified set of functional synaptic properties. In addition, some statements-particularly those linking coupling mode to "higher-order neocortical functions"-appear broader than what is directly supported by the experiments and should be tempered or more precisely scoped.

Below, I list several topics that could help better frame the main findings of the present study and clarify how it relates to previously published work.

We would like to thank the reviewer for the comprehensive and detailed assessment of our manuscript and his/her overall positive evaluation. We have addressed all of the reviewer's points. We expanded the model description and discussion, and slightly toned down our conclusion.

(1) The authors use EGTA sensitivity of EPSCs (together with additional metrics) to argue that S1 and PFC synapses differ in Ca2+ channel-release sensor coupling. While this is a plausible interpretation, EGTA effects are not uniquely determined by coupling distance and can also reflect differences in Ca2+ entry kinetics, action potential waveform, endogenous buffering/extrusion, or release-sensor/vesicle state. The authors use a constrained modeling approach, but the rationale for the different constraint sets is not fully clear from the current description. It would be helpful to expand and clarify the Methods section to explain how these constraints were defined, justified, and applied (and how alternative constraint choices would affect the results). In this context, the Abstract's broader claim that the study "reveals microdomain coupling as a presynaptic structure-function correlate of higher-order neocortical functions" appears overstated. Given the well-known diversity of cortical synapses even within a single region (e.g., synapses onto different interneuron subclasses or different PN cell types, extracortical sources like thalamus), the authors should clarify the intended scope: is the conclusion meant to apply broadly across synapse classes in S1 and PFC, or only to the specific connection type(s) examined here?

We would like to thank the reviewer from pointing out that our description fell a bit short, in particular with respect to the interpretation of the EGTA effects. We addressed the points as follows in the revised manuscript: We discussed the interpretation of EGTA effects in more detail. We toned down the concluding statement in the last sentence of the Abstract.

“Differences in the sensitivity of release to low to moderate concentrations of EGTA (≤ 30 mM) are a standard indicator of differences in the coupling distance (e.g. Adler et al., 1991; Bucurenciu et al., 2008; reviewed in Eggermann et al., 2012; Vyleta and Jonas, 2014; Kusch et al., 2018; Bornschein et al., 2019b). pN is determined by the size of the Ca2+ signal at the release sensor and the binding kinetics and affinity of the sensor. The former in turn is determined by the details of the Ca2+ influx and the diffusional coupling distance between the VGCCs and the sensor. The similarity of Ca2+ signals between synapses in PFC and S1 (Figure 4) indicates that Ca2+ influx is similar between boutons, although more subtle differences in the influx kinetics may have remained undetected in these volume-averaged signals. Regarding sensor affinity, results in a previous study indicate that differences in EGTA sensitivity show differences in coupling rather than sensor affinity even if kon of the sensor and its affinity should differ as much as ten-fold, which appears to be an unlikely scenario given that even the two major isoforms of Synaptotagmin that trigger synchronous release differ by less than a factor of three to four in their affinity (Bollmann et al., 2000; Schneggenburger and Neher, 2000; Bornschein et al., 2025). Finally, the increase in the PPR induced by the application of Cd2+ further supports our conclusion of microdomain coupling in the PFC synapses (Scimemi and Diamond, 2012).”

“They suggest that microdomain coupling in pyramidal neuron synapses could be a presynaptic structure-function correlate of higher order neocortical functions.”

(2) The chelator logic is sound in principle, but the Discussion should more explicitly acknowledge standard caveats and alternative explanations. The authors partly address this by including presynaptic Ca2+ imaging and modeling, yet it would help to explain more clearly how the combination of (i) chelator sensitivity, (ii) presynaptic Ca2+ signals, and (iii) model constraints rules out-or substantially reduces the likelihood of-changes in AP waveform, Ca2+ influx kinetics, buffering/extrusion, or sensor/vesicle state as the primary drivers. In addition, recent hypotheses emphasizing vesicle priming and/or release-site occupancy as contributors to apparent EGTA sensitivity should be discussed as a complementary or alternative interpretation.

Please see above the first part of the discussion to point one.

(3) A substantial portion of the S1 comparison appears to rely on previously published datasets. This should be made unambiguous in the Results and Methods, and it would be helpful to summarize this clearly (e.g., in a table indicating which figures/analyses use new data versus reanalysis of published data). If this information is already present, it should be highlighted more prominently.

Please excuse us for not having made it clearer which data had already been published. Only the paired recordings from L5PN-L5PN connections in S1 were obtained in previous studies and partially reanalyzed. Paired recordings on the same synapses in PFC as well as all recordings from L2/3-L5PN connections in PFC and S1 were obtained in the present study. At your suggestion, we have added Table 1 highlighting which data and associated figures are from this study and which were acquired in previous studies (Bornschein et al., Cell Rep. 2019; Bornschein et al., Front. Syn. Neurosci. 2019).

(4) The modeling is informative, but the choice of a specific VGCC-release-site geometry and channel arrangement is not sufficiently justified. The manuscript adopts a particular spatial configuration, yet the rationale for selecting this geometry, rather than other plausible architectures discussed in the literature, is not clearly explained, nor is it meaningfully revisited in the Discussion. The authors should justify why the same organization is assumed across two distinct cortical areas and, ideally, include (or at a minimum discuss) a sensitivity analysis showing how key inferences (e.g., coupling distance and channel number) depend on the assumed geometry.

We extended the discussion of why a ring-like structure of VGCCs was assumed in the model.

“The microdomain was assumed to be formed by a ring-like structure of VGCCs around a vesicle (Figure 5D). This topography was chosen because such a microdomain was found to best predict the experimental data of transmitter release from PNs in young S1 (Bornschein et al., 2019b). Other previously described distributions of VGCCs suitable to reproduce release data cover random distributions of VGCCs (Scimemi and Diamond, 2012), VGCC clusters (Meinrenken et al., 2002; Nakamura et al., 2015), and exclusion zones (Keller et al., 2015). In the early S1, all of these models predicted a higher EGTA sensitivity of the microdomain, however, these models provided a poorer fit to the full set of the experimental data than the ring-like structure (Bornschein et al., 2019b). Since the experimental data from PNs in the mature PFC were similar to those in young S1, these other microdomain models were not tested explicitly here.”

(5) The calcium imaging data are valuable, but given the diversity of synapses within each cortical layer, it is not clear that imaged boutons can be confidently assigned to the specific connection types being interrogated electrophysiologically. A substantial fraction of boutons likely corresponds to different postsynaptic targets (including interneurons and distinct pyramidal-cell classes), and this heterogeneity could complicate interpretation. This limitation should be discussed explicitly

Excitatory pyramidal cells make up 80-85% of cortical neurons, with the highest density in layer 5 (Keller et al., Front. Neuroanat. 2018). In the somatosensory cortex, inhibitory synapses account for only about 10% (Santuy et al., Brain Struct. Funct. 2018). We imaged a large number of presynaptic boutons within layer 5 (about 10 boutons per cell, in total 85 boutons in PFC and 100 boutons in S1, numbers of boutons were now included in Figure 4). In this respect, the impact of inhibitory synapses is minor. Since connectivity between neighboring PNs in layer 5A is high (Feldmeyer, Front. Neuroanat. 2012), we assume that a large proportion of the imaged boutons target neighbouring L5PNs. We added a sentence on potential postsynaptic targets in the results section.

“The imaged presynaptic boutons most likely connect to neighboring pyramidal cells, as connectivity between L5PNs in layer 5A is high (Feldmeyer, 2012). Nevertheless, a small proportion of other postsynaptic targets, such as interneurons, cannot be ruled out.”

(6) In unitary connections, the authors assess EGTA effects alongside other functional parameters (strength, delay, short-term plasticity), which is a major strength. However, for L2/3 to L5 connections, it appears that EGTA sensitivity was tested primarily using extracellular stimulation. Given anatomical and circuit differences between PFC and S1, extracellular stimulation may recruit different synapse populations across regions, potentially confounding regional comparisons of EGTA sensitivity. This limitation should be acknowledged explicitly. While I am not requesting technically demanding L2/3↔L5 paired recordings in S1, the possibility that different synapse identities are being sampled should be treated as a meaningful source of uncertainty. The Discussion would also benefit from placing the magnitude of EGTA effects in the context of prior "loose coupling" literature, where comparatively large EGTA effects have been reported in some systems. In addition, the reported difference between adult PFC EGTA effects and S1 inhibition appears small (on the order of <10%) and should be interpreted cautiously, especially given that PFC and S1 mature on different timelines and P21-P26 is unlikely to reflect a mature PFC circuit state. The adult cohort (P90-P100) is therefore important, but the age mismatch complicates PFC-S1 comparisons; ideally, S1 should be assessed at matched ages, or this limitation should be discussed explicitly. Finally, for statistical robustness, in panel D of Figure 2, were the comparisons corrected for multiple testing to control Type I error?

EGTA sensitivity was examined in PFC and in S1, for two connections in each region - using paired recordings for L5PN-L5PN connections and using extracellular stimulation for L2/3-L5PN connections. The L5PN-L5PN data from S1 were collected in an earlier study (Bornschein et al., Cell Rep. 2019; Bornschein et al., Front. Syn. Neurosci. 2019), have now been reanalyzed for the test period between 20 and 30 min, and included in Figure 2B for the sake of consistency (see Table 1). To emphasize this point, despite stimulating different input synapses with different stimulation methods, we obtained similar results in the respective brain regions. This suggests that the EGTA sensitivity observed in the investigated PFC connections is not a solely synapse-specific property.

It is difficult to compare the absolute EGTA sensitivities from different synapses from different publications, since EGTA effects do not depend exclusively on the coupling distance, as the reviewer also noted in point 1. They are, among other factors, influenced by the Ca2+ sensitivity of the release machinery, which differs between our Syt1-expressing cortical synapses and Syt2-expressing synapses in other brain regions (Schneggenburger et al., Nature, 2000; Bollmann et al., Science, 2000; Bornschein et al., Science 2025), such as the calyx of Held or the cerebellar basket to Purkinje cell synapse. Furthermore, direct patching and loading of the presynaptic bouton with EGTA - as feasible at the calyx of Held and other large synapses - results in higher effective EGTA concentrations compared to somatic loading of presynaptic terminals, despite identical pipette concentrations. The buffer-AM method introduces additional uncertainty regarding the effective intra-bouton EGTA concentration, since the loading efficacy has to be estimated. Thus, although differences in EGTA sensitivity primarily show differences in coupling distances, the comparison of absolute values between different publications is difficult. Consistently, data-constrained models are used to estimate the coupling topography and to compare these topographies rather than comparing the absolute EGTA effects (e.g. Buccurenciu et al., Neuron, 2008; Vyleta and Jonas, Science, 2014; Bornschein et al., Cell Rep., 2019; Chen et al., Neuron, 2024; Bornschein et al., Science, 2025).

We include a note on this in the discussion.

“Thus, although the absolute EGTA sensitivity is influenced by different factors, which necessitates data-constrained models for quantitative comparisons, the general sensitivity of release to EGTA indicates loose coupling.”

In mouse neocortex postnatal maturation in S1 and PFC follows the same time course. Kroon et al. (Sci. Rep. 2019) reported that maturation of dendritic morphology and intrinsic properties of pyramidal neurons occurs within the first two weeks after birth, now cited in the discussion. Therefore, it is unlikely that the EGTA effect in PFC is due to a delayed maturation. The difference in EGTA sensitivity between PFC at P90-100 and S1 at P21-26 is indeed small but significant (P=0.009, Mann-Whitney-U rank sum test).

“Since postnatal development follows the same time-course in mouse PFC and S1 and occurs predominantly within the first two weeks after birth, (…) (Kroon et al., 2019).”

Thank you for the advice concerning statistical robustness. We replaced the Mann-Whitney-U rank sum test in Figure 2D by a one-way ANOVA and performed a Holm-Sidak post-hoc test correcting for multiple comparisons. Similarly, ANOVA was used to compare more than two groups in Figures 1K and 2F. We changed the corresponding P values and tests in the figure legends and added the performed post-hoc tests in the methods section.

“For multiple comparisons post-hoc testing was performed with the Holm-Sidak (one-way ANOVA) or Dunn´s method (ANOVA on ranks).”

(7) Alterations in initial release probability are often associated with changes in short-term plasticity. In the present manuscript, the authors report similar initial release probability at PFC and S1 synapses, yet observe differences in short-term plasticity profiles. The mechanistic basis for this apparent dissociation is not addressed and should be discussed explicitly, including potential explanations.

Various other factors besides pN can influence short-term plasticity, that are the coupling distance, the number of occupied release sites (Nocc), the replenishment of Nocc or the recruitment of newly formed Nocc as well as the expression of endogenous Ca2+ buffers (Blatow et al., Neuron 2003; Felmy et al., Neuron 2003; Matveev et al., Biophys. J. 2004; Neher, Cell Calcium 1998; Regehr, CSH Perp. Biol. 2012) or fascilitation sensors (Turecek & Regehr, J. Neurosci. 2018; Shin et al., eLife 2025). Traditionally, pN had been assumed to have a major impact on short-term plasticity (STP) which is indeed the case at low replenishment rates (e.g. Feldmeyer and Radnikow, J. Physiol. 2009; Zucker and Regehr, Ann. Rev. Physiol. 2002). But at several synapses very fast replenishment rates have been described driving a progressive overfilling of the initial RRP and increasing Nocc above baseline levels (Brachtendorf et al., Front. Cell. Neurosci. 2015; Doussau et al., eLife 2017; Miki et al., Neuron 2016; Valera et al., J. Neurosci. 2012) making replenishment the stronger determinant of STP.

Additionally, the size and organization of sub-pools from which vesicle recruitment and release occurs affects the speed and reliability of vesicular release. In our previous study on L5PN-L5PN connections in S1 we found that developmental tightening of CDs was associated with an increase in PPR without altering pN (Bornschein et al., Cell Rep. 2019). We could show that the maturation of a replenishment pool during postnatal development increases vesicle recruitment and reliability thereby affecting STP (Bornschein et al., Front. Syn. Neurosci. 2019).

We discussed this in the revised manuscript.

“Classically, pN was considered as the major determinant of short-term plasticity (e.g. reviewed in Zucker and Regehr, 2002; Feldmeyer and Radnikow, 2009). More recently other factors, including the number of occupied release sites, their replenishment or an increase in their occupancy, or the expression of endogenous Ca2+ buffers have been considered as more important determinants of short-term plasticity (Rozov et al., 2001; Blatow et al., 2003; Felmy et al., 2003; Matveev et al., 2004; Bornschein et al., 2013; Miki et al., 2016; Doussau et al., 2017; Jackman and Regehr, 2017; Neher and Brose, 2018). (…)”

Short-term plasticity changes during postnatal development at different cortical PN connections without alterations in pN (Reyes and Sakmann, 1999; Bornschein et al., 2019a). For L5PN-to-L5PN connections these differences were found to result from the maturation of an intermediate replenishment vesicle pool (Bornschein et al., 2019).

(8) There are multiple instances where the text appears to cite non-existent or misnumbered figure panels (e.g., references to "Figure 4G-I / 4J" when the relevant material appears elsewhere). These should be corrected throughout, as they currently reduce readability and confidence.

We apologize for the misnumbering which originated from a previous version of this manuscript. Figure 4G-J is actually Figure 5A-D. We corrected the references to Figure 5 in the methods section.

(9) The Methods describe P21-P26 animals, whereas the Results include older cohorts (e.g., P90-P100) and additional regions (e.g., mPFC). The Methods should be updated so that all cohorts and regions analyzed in the Results are fully described.

Thank you for thoroughly reading the methods. We added the missing cohort (P90-100) and brain region (mPFC) to the method section.

“C57BL/6J mice at P21-26 and P90-100 of either sex were decapitated under deep Isoflurane (Curamed) inhalation anaesthesia. (…) Coronal neocortical slices (150-250 μm thick) were cut from the lateral PFC, medial PFC (mPFC) or S1 region (Figure 1A) with a vibratome (HM 650 V, Microm).”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

These are mostly points for discussion; there is no need for additional experiments.

(1) Discuss potential effects of (re-)patching on presynaptic physiology and the "control" time course in Figure 2A.

The repatching strategy is a strength, but it also introduces opportunities for physiological drift (dialysis effects, changes in access resistance, altered excitability, and the switch to presynaptic stimulation in whole-cell after repatching). Please discuss (and, if already quantified, briefly report) why EPSC amplitudes appear to increase in the control time course after repatching (as seen in Figure 2A). Even a short explanation (e.g., run-up after whole-cell access, recovery from on-cell stimulation, washout of endogenous buffering, improved spike waveform reliability, etc.) plus reassurance that baseline stationarity criteria were met would strengthen confidence in the repatching-based inference.

Baseline recordings were usually performed with the presynaptic neuron in cell-attached mode. In this configuration intracellular ion concentration, second messenger systems as well as cell-specific resting membrane potential stay essentially unaffected. When switching to whole-cell mode after re-patching of the presynaptic cell, the pipette solution determines the intracellular environment, which may also affect second messenger systems and mobile endogenous buffers will be washed out. Although layer 5 pyramidal neurons do not express relevant concentrations of these mobile buffers (Helmchen et al., Biophys. J., 1996; Tran and Stricker, Biophys. J., 2018; Bornschein et al., Cell Rep., 2019), this might have contributed to the moderate and temporary run-up after whole-cell access to presynaptic cells.

In postsynaptic neurons, we also routinely controlled the stability of Rs and Ileak during our prolonged measurements. In the controls, we found a median initial increase in relative EPSC amplitudes to 1.17 (1.02-1.38) between 0 and 10 min after the presynaptic whole-cell access was established, which correlated with a temporary decline in Rs in 3 out of 6 recordings. EPSC amplitudes returned to their baseline values after 20 min at the latest (1.02, 0.75-1.20). We added potential reasons for the temporary amplitude increase in the controls in the results section.

“In control recordings a temporary initial increase in EPSC amplitudes after whole-cell access to the presynaptic neuron was evident. Although L5PNs do not express large concentrations of mobile buffers (Helmchen et al., 1996; Tran and Stricker, 2018; Bornschein et al., 2019b), their wash-out might have contributed to the temporary run-up. On the other hand, run-up correlated with a temporary decline in Rs in some recordings (3 out of 6). Run-up effects normalized after 20 min at the latest (1.02, 0.75-1.20).”

(2) Clarify interpretation/robustness of MPFA-derived "N" given the large range, and discuss uncertainty from using only three [Ca2+]e conditions for L2/3-L5PN MPFA.

The manuscript reports that "N" is markedly smaller in PFC than S1 (median ~2.1 vs 8), but the S1 range is very broad (3-19).

(a) Briefly discuss whether/why such a wide N range is expected and how it should be interpreted (binomial "N" vs anatomical release sites; sensitivity to CV assumptions; potential dependence on connection geometry, bouton number, dendritic filtering, etc.).

(b) Add a short statement on uncertainty/identifiability when fitting MPFA with only three conditions for L2/3-L5PN (e.g., whether confidence intervals/bootstraps were examined; how stable q and N are to small changes in the variance estimates). Even a qualitative note would help readers judge how much weight to put on the absolute N estimates versus the overall cross-area trend.

(a) In our previous study on this connection we determined a wide range for N (8, 3-19) even though we used four extracellular Ca2+ concentrations in MPFA (Bornschein et al., Cell Rep. 2019). The binomial parameter N can be considered to represent the number of release sites, including empty release sites (see Brachtendorf et al., Front. Cell. Neurosci. 2025). The range of release sites is likely to reflect the variability in the number of anatomical synaptic contacts ranging from 1 to 6 for these synapses (Frick et al., Cereb. Cortex 2008). Since 1 to 3 active zones/release sites per synaptic contact appear to be typical for small cortical synapses (e.g. Xu-Friedman et al., J. Neurosci. 2001), this results in a wide range of 1-18 release sites per connection.

In our previous study (Bornschein et al., Cell Rep. 2019), we also investigated the effects of different values of CV1 and CV2 by repeating the MPFA fitting procedures for different combinations of CV1 and CV2 ranging from 0.1 to 1 each. We have quantified a deviation of ≤10% in the estimates of vesicular release probability from the typically used CV values of 0.3 across a wide range of CV value combinations (see also Schmidt et al., Curr. Biol. 2013). We have added a note regarding CV sensitivity in the methods section.

“For CV assumptions that deviate from the standard value of 0.3, deviations in the calculated pN values of less than 10% are to be expected (Schmidt et al., 2013; Bornschein et al., 2019b).”

(b) The three Ca2+ concentrations we used for MPFA resulted in a low (<0.5), a medium (~0.5) and a large (>0.5) pN condition. With this, a parabola is uniquely determined by three parameters. To further ensure the reliability of the parameters determined by MPFA, we compared them to values estimated from EPSC amplitudes (EPSC = N pN q; PFC, 6 pA; S1, 48 pA) and failure rates (F = (1-pN)^N; PFC, 0.25, S1, 0.0001), which yielded values similar to those from MPFA (EPSCs in PFC: 8 pA, 5-15 pA, and S1: 29 pA, 18-53 pA; failure rates in PFC: 0.16, 0.08-0.28, and S1: 0, 0-0.03; see original manuscript).

To further support this, we have now conducted additional experiments using four Ca2+ concentrations. The results are consistent with those from the experiments using three concentrations. The additional experiments are shown for review purposes in Figure R1. The novel pN data are included in the summary of pN values (now n=6) in the results section and in Figure 3F.

Additionally, we performed bootstrap analyses with 10,000 replicates which were generated with replacement from the original data sample. Distributions of bootstrap 25% trimmed means showed a clear separation for N and q between PFC and S1 but not for pN. The bootstrap results are shown for review purposes in Author response image 2 (cf. point 5 of Reviewer#2).

(3) Broaden the discussion, e.g. by linking to nanodomain/microdomain coupling as a general strategy for stimulus encoding, including sensory periphery examples.

The work will resonate beyond the cortex if the authors explicitly connect their findings to broader principles: how the spatial coupling regime shapes the transfer function between Ca2+ entry and vesicle fusion, thereby tuning reliability, timing, and dynamic range. Requested addition: Please consider adding a short subsection discussing analogous implementations in the sensory periphery, especially ribbon synapses of cochlear inner hair cells and rod photoreceptors, where nanodomain coupling has been discussed as a key determinant of encoding and release dynamics. Also, citing relevant work such as that by Scimemi and Diamond 2012 would further strengthen the paper.

We agree that the work by Scimemi and Diamond (J.Neurosci. 2012) is important and we cited and discussed their work in several of our previous publications. We now also included the paper in the revised version of the present manuscript. As requested, we also included a discussion on findings from ribbon type synapses and also from the neuromuscular junction.

“Nanodomain coupling was also found in the peripheral nervous system, in particular at retinal (Singer and Diamond, 2003; Jarsky et al., 2010) and auditory (Moser and Beutner, 2000; Brandt et al., 2005) ribbon-type synapses and at the neuromuscular junction (Harlow et al., 2001; Shahrezaei et al., 2006). These synapses have highly specialized properties and appear to be optimized for very reliable transmission and, in the case of ribbon synapses, also for high-frequency coding of sensory information (reviewed in Matthews and Fuchs, 2010; Eggermann et al., 2012). Thus, it appears that synapses in the sensory pathways, in particular those engaged in reliable high-frequency coding of sensory information, both in the periphery and in the lower processing stages of the CNS, up to primary sensory cortices, operate with nanodomain coupling. In the executing motor pathway, the neuromuscular junction uses nanodomain coupling and, as recent results from our group suggest, also PNs in the primary motor cortex (Yarim et al., in preparation). It is tempting to speculate that complete loops from or to the primary cortices to their peripheral target organs operate with nanodomains. Microdomain coupling, on the other hand, appears to come into play only if integration of information from multiple sources and plasticity are the main focus, as at certain synapses in PFC (this study) or hippocampus (Vyleta and Jonas, 2014).”

“The microdomain was assumed to be formed by a ring-like structure of VGCCs around a vesicle (Figure 5D). This topography was chosen because such a microdomain was found to best predict the experimental data of transmitter release from PNs in young S1 (Bornschein et al., 2019b). Other previously described distributions of VGCCs suitable to reproduce release data cover random distributions of VGCCs (Scimemi and Diamond, 2012), VGCC clusters (Meinrenken et al., 2002; Nakamura et al., 2015; Rebola et al., 2019), and exclusion zones (Keller et al., 2015; Rebola et al., 2019). In the early S1, all of these models predicted a higher EGTA sensitivity of the microdomain, however, these models provided a poorer fit to the full set of the experimental data than the ring-like structure (Bornschein et al., 2019b). Since the experimental data from PNs in the mature PFC were similar to those in young S1, these other microdomain models were not tested explicitly here.”

“(…) Finally, the increase in the PPR induced by the application of Cd2+ further supports our conclusion of microdomain coupling in the PFC synapses (Scimemi and Diamond, 2012).”

(4) Address limitations of basal/resting Ca2+ estimates and make explicit that measured Ca2+ signals are volume-averaged (not microdomain) readouts.

(a) The reported basal [Ca2+]i values are in the ~tens of nM range. Given the stated in vitro KD for Fluo-5F in the authors' pipette solution (439 nM), the resting estimates are far below KD; this does not invalidate the approach, but it does warrant a brief discussion of sensitivity/uncertainty (influence of Rmin estimation, background subtraction, and how errors propagate into basal [Ca2+]i). Repeating experiments is not necessary-just clearer framing of limitations.

(b) Please also emphasize more prominently (ideally in Results and/or Discussion) that the bouton signals are volume averaged and therefore do not directly report calcium microdomains at active zones or nanodomains at release sensors. The Methods already state this point; echoing it in the main text would prevent over-interpretation by readers.

(a) We agree that Fluo5F is less suitable for determining absolute basal calcium levels. In a previous study (Bornschein et al., Science 2025) we determined the basal Ca2+ concentration with OGB1 (KD=166 nM; basal [Ca2+]i=44 nM, 24-58 nM, n=43 boutons from 10 cells) and observed no significant difference to basal [Ca2+]i values determined with Fluo5F despite the KD of 439 nM (31 nM, 16-54 nM, 14 boutons from 3 cells; P=0.204, MWU; data not published). We added this limitation to the results section and swapped Figure panels 4E and F for confluence. The calibration curve of Fluo5F as well as the comparison to basal [Ca2+]i values determined with OGB1have been included in Figure S4.

“The quantification of absolute basal [Ca2+]i was limited by the KD of Fluo5F (439 nM), which slightly underestimated basal [Ca2+]i values in comparison to quantification with OGB1 (KD=166 nM, Figure S4). Nevertheless, relative comparison of basal [Ca2+]i yielded no significant differences between PFC (30 nM, 21-34 nM) and S1 (22 nM, 13-38 nM; Figure 4F).”

(b) In the results section, we have now emphasized that volume-averaged Ca2+ signals were measured.

“We performed dual-dye two-photon Ca2+ imaging (Sabatini et al., 2002) to quantify volume-averaged Ca2+ signals at presumed presynaptic boutons located on axon collaterals of L5PNs in PFC and in S1.”

Reviewer #2 (Recommendations for the authors):

(1) For a meaningful comparison, recordings from the PFC and the S1 cortex of the same animals should be undertaken. Additionally, I suggest performing additional experiments regarding the different cell types of L5 pyramidal cells in layer 5a.

We performed new experiments to determine EGTA sensitivity in PFC and S1 from the same animal. The results from these experiments agree with the previous results. They are included in the results section , in Figure 2C-F and in Figure S3A-C.

Additionally, we extended the discussion on the examined cell types. For S1 cortex we refer in more detail to our previous work, where we described in depth where and under consideration of which criteria our recordings were established and that based on these criteria we recorded from pyramidal neurons in layer 5A in S1 (Bornschein et al., Cell Rep. 2019; Bornschein et al. Front. Synapt. Neurosci. 2019; Bornschein et al., Science 2025). Within layer 5A, we did not attempt to further distinguish between types of pyramidal neurons. We include this in the methods section.

“Patch-clamp recordings from L5PNs located in the upper layer 5 (L5A in S1) were established according to the criteria described in detail in our previous work on this connection in S1 (Bornschein et al., 2019b; Bornschein et al., 2025). Presynaptic neurons were stimulated extracellularly in upper layer 2/3 (L2/3-L5PN connections) straight above the patched L5PN or in on-cell mode in L5A right next to the postsynaptic cell (L5PN-L5PN connections; Figure 1).”

For the recordings in PFC and heterogeneity in pyramidal neuron types we refer to our detailed response to the point 3 of Reviewer 2. There we also discuss that the heterogeneity in morphology and spiking patterns is probably not reflected on the synaptic level. We would also like to emphasize that the type of experiments we perform with paired recordings and long-lasting patch-clamp measurements is not suitable to differentiate between subpopulations of pyramidal neurons. This would require successful recordings form several tens of different pyramidal neurons, which is not feasible in our type of experiment. We discuss this limitation of the discussion.

(2) The authors need to comment in depth on their MPFA data, and if feasibl,e perform additional experiments.

Concerning the robustness of quantification of synaptic parameters by MPFA, we refer to our comments on point 2b of the recommendations for the authors to Reviewer 1. Additionally, we performed new MPFA experiments with four extracellular Ca2+ concentrations that agree with our results with three Ca2+ concentrations.

(3) The statistical analysis should be revised and a test for the normality of data distribution should be implemented.

A test for normal distribution (Shapiro-Wilk test) has already been described in the methods section in the previous version of this manuscript.

(3) Figure 1A is somewhat misleading because it could suggest that the authors have performed dual recordings in identified PFC pyramidal cells.

We added “L2/3 or L5” to the stimulation panel of Figure 1A to illustrate that we stimulated either extracellularly in L2/3 or L5PNs directly via the patch pipette.

(4) Is the relative variance of the mean EPSC amplitude and latency between connections larger in the PFC connections than in S1 cortex? This could indicate a variability in cell types.

The relative variance of EPSC amplitudes calculated as median absolute deviation (MAD) was 0.46 in PFC and 0.50 in S1 arguing against differences in the variability in cell types. The larger variability in delays expressed as SDDelay is the result of the larger coupling distance in PFC compared to S1 (Bullmann et al., J. Neurosci. 2024). Consequently, also the relative MAD is larger (0.89) in PFC compared to S1 (0.14) and is therefore not able to detect differences in the variability of recorded cell types.

(5) Reyes and Sakmann (1999) have previously described differences for L2/3-L5b and L5b-L5b synaptic connections in S1 cortex at different developmental stages. This paper needs to be cited as it is highly relevant to this study.

Reyes and Sakmann (J. Neurosci. 1999) reported layer-specific differences in short-term plasticity in young sensorimotor cortex which disappeared as maturation progressed and short-term plasticity increased. In a previous study (Bornschein et al., Front. Syn. Neurosci. 2019) we also described a developmentally driven increase in short-term plasticity caused by the maturation of vesicle pools. In the present study we used mature animals and would therefore not expect layer-specific differences neither in S1 nor in PFC since the time course of postnatal maturation was described to be comparable in both neocortical circuits (Kroon et al., Sci. Rep. 2019).

We discussed this paper in the context of developmental changes in short-term plasticity.

“Short-term plasticity changes during postnatal development at different cortical PN connections without alterations in pN (Reyes and Sakmann, 1999; Bornschein et al., 2019a). For L5PN-to-L5PN connections these differences were found to result from the maturation of an intermediate replenishment vesicle pool (Bornschein et al., 2019a). Such pool maturation may also underlie the elimination of layer-specific differences in short-term plasticity between L2/3-L5B and L5B-L5B synaptic connections that were evident in young rats but eliminated during the first weeks of postnatal development (Reyes and Sakmann, 1999). Since postnatal development follows the same time course in mouse PFC and S1 and occurs predominantly within the first two weeks after birth, significant layer-specific differences in PN synapses are unlikely in both areas in our experimental time window (Kroon et al., 2019). Consistently, we found similar PPRs at L2/3-L5PN synapses and L5PN-L5PN synapses in both areas, with facilitation in PFC and depression in S1, irrespective of the presynaptic PN synapse type.

(6) Regarding the point of loose or tight Ca2+ channel coupling: Could some of the differences result from differences in the presynaptic Ca2+ channel complement? Please comment.

This can be excluded. Cav2.1 and Cav2.2 are the main channels gating release at PN synapses. The gating kinetics of these channels are very similar and they only differ somewhat in their peak current amplitude (Bornschein et al., Cell Rep. 2019, Figure 4). Since the number of open channels is a fit parameter in our simulations there would only be an effect on the estimate of the number of channels gating release but not for the estimate of the coupling distance. This is all the more true since the EGTA effect depends on the diffusion distance rather than on the gating kinetics. These considerations will also hold for Cav2.3 channels, which have slower closing kinetics, but anyway play only a very minor role for triggering release.

Version 2

Abstract

In the mature neocortex, functionally distinct areas are built by the same archetypes of neurons, but depending on the area, these neurons and their synapses are engaged in very different functions, ranging from lower order processing of sensory information to higher order associations and cognitive functions. We found significant differences in the functional presynaptic nanoarchitectures of the same types of pyramidal neuron synapses, depending on whether they are located in the prefrontal cortex (PFC) or in the primary somatosensory cortex (S1). Synapses in PFC operated with loose microdomain coupling as opposed to tight nanodomain coupling in S1. These differences were associated with significant differences in synaptic timing, efficacy and plasticity between areas. Our data suggest that the mature neocortex uses tuning of synaptic topographies to specialize seemingly identical types of neurons for their required function. They suggest that microdomain coupling in pyramidal neuron synapses could be a presynaptic structure-function correlate of higher order neocortical functions.

Introduction

The neo- or isocortex is a morphologically homogeneous brain region that covers highly diverse functions within its specialized areas, ranging from early sensory processing and motor control up to higher order associations, cognition and consciousness (Douglas and Martin, 2004; Harris and Shepherd, 2015). Notably, all of these diverse functions are executed by the same archetypes of neurons and synapses. However, depending on the area, they are engaged to different degrees in lower or higher order processing. Whether area-specific functional differences in the mature neocortex are associated with or even arise from differences in the functional presynaptic nanoarchitecture of the same principal types of synapses is currently unclear.

At presynaptic active zones, the coupling distance between voltage-gated Ca2+ channels (VGCCs) and transmitter-filled synaptic vesicles (SVs) is a major determinant of key synaptic properties, including speed, efficacy and reliability of synaptic transmission (Rozov et al., 2001; Fedchyshyn and Wang, 2005; Bucurenciu et al., 2008; Baur et al., 2015; Bornschein et al., 2019b; Chen et al., 2024), and also short-term plasticity, although the latter can be obscured by vesicle replenishment (Miki et al., 2016; Doussau et al., 2017; Bornschein et al., 2019a; Lin et al., 2022). Studies at different synapses in different parts of the brain suggest that excitatory synapses engaged in reliable information transfer switch from loose microdomain to tight nanodomain coupling during development (Fedchyshyn and Wang, 2005; Baur et al., 2015; Nakamura et al., 2015; Bornschein et al., 2019b). This includes synapses between layer 5 pyramidal neurons (L5PNs) in the primary somatosensory cortex (S1), an area of early sensory processing (Bornschein et al., 2019b). On the other hand, loose microdomain coupling was found in the mature brain to date only at a highly plastic hippocampal synapse and has been suggested to provide a molecular framework for presynaptic plasticity (Vyleta and Jonas, 2014). However, if microdomain coupling also represents a synaptic correlate of higher order neocortical function or if it restricted to specific and highly plastic synapses in the hippocampus is currently unclear. Therefore, here we tested the hypothesis that loose microdomain coupling is the synaptic correlate of higher order functions of the mature neocortex, whereas tight nanodomain coupling is the synaptic correlate of earlier processing stages. Our data show that in the mature cortex the same types of pyramidal neuron synapses operate with loose microdomain coupling if they are located in the prefrontal cortex (PFC), whereas in S1 they operate with tight nanodomain coupling. Thus, our results corroborate the hypothesis and suggest microdomain coupling as a synaptic correlate of higher order function in the neocortex.

Results

Synaptic efficacy, reliability and short-term plasticity differ between PFC and S1

We focused on the lateral prefrontal cortex (PFC) and medial prefrontal cortex (mPFC) as areas of higher order processing and on primary somatosensory cortex (S1) as an area of lower order processing. Within these areas we investigated excitatory synapses between pyramidal neurons (PNs), which are the principal building blocks of the neocortex. We first analyzed the properties of synapses connecting neighboring layer 5 PNs (L5PNs) in mature PFC (P21-26) by performing paired whole-cell patch-clamp recordings and compared their properties to those we previously reported for the same synapses and age in S1 (Bornschein et al., 2019a; Bornschein et al., 2019b) (Figure 1A, Table 1). S1 data from both previous studies were combined and additional analyses of synaptic delays, PPRs and success rates were performed. We found that the efficacy of transmission, i.e. the size of a unitary EPSC evoked by a single action potential (Rozov et al., 2001), was significantly lower in PFC (8 pA, 5-15 pA) than in S1 (29 pA, 18-53 pA), whereas the synaptic failure rates were significantly higher in PFC (0.16, 0.08-0.28) than in S1 (0, 0-0.03; Figure 1B-D). In addition, we found that the synaptic delays (2.46 ms, 2.20-2.94 ms) as well as their standard deviation (SDDelay: 0.60 ms, 0.49-0.81 ms) were significantly larger in PFC than in S1 (delay: 1.99 ms, 1.67-2.25 ms; SDDelay: 0.24 ms, 1.14-0.37 ms; Figure 1E, F), indicating that transmitter release is less tightly coupled to the time of the action potential in PFC than in S1 (Bullmann et al., 2024). Together these data show that synaptic speed, efficacy and reliability during single action potentials are lower in PFC than at the same synapses in S1.

Differences in synaptic transmission and plasticity between PFC and S1 synapses.

(A)Schematic representation of the mature mouse brain (top) and the positions at which coronal sections were made from PFC (orange; middle) or S1 (blue; bottom). Recordings were made in acute slices from L5PN-L5PN pairs or from L2/3-L5PN connections. (B)EPSCs (bottom) recorded from postsynaptic L5PNs after evoking action currents in the presynaptic L5PNs (top, inset: 3fold magnification) in PFC (left) or S1 (right), individual recordings in gray, averages in black). (C-F) Summary of basic synaptic properties: EPSC amplitudes (C), failure rates (D), synaptic delays (E) and their SD values (F) in PFC and S1 (medians ± IQRs, means as dashed lines, whiskers represent 10th and 90th percentiles, dots indicate outliers, numbers of cell pairs in brackets; ***P<0.001, Mann-Whitney U test (MWU)). Data from S1 are from ref. (Bornschein et al., 2019b). (G) EPSCs recorded from pairs of L5PNs at 50 Hz intervals in PFC (top) and S1 (bottom), presynaptic stimulations are omitted, averages from at least 10 individual recordings). (H, I) Summary of PPRs (H; ***P<0.001, MWU) and success rates (I) following the 1st (1_x) and 2nd stimulation (x_1; ***P<0.001, P=0.987, Wilcoxon signed rank test (WSR)). S1 data are from ref. (Bornschein et al., 2019a). (J) As in (G) but for L2/3-L5 synapses (bottom, averages from at least 15 individual recordings, stimulation artefacts were blanked for clarity, arrowheads denote the time points of extracellular stimulation). (K) Boxplots summarizing PPR data from lateral PFC (PFC), medial PFC (mPFC), older PFC (P90-100) and in S1 (***P<0.001, n.s., not significant, ANOVA on ranks). (L) Comparison of SDs of the synaptic delays at L2/3-L5 synapses between PFC and S1 (***P<0.001, MWU).

When pairs of action potentials were applied at high frequency (Figure 1G-I), differences in the synaptic success rates were associated with a significant difference in the paired pulse ratios (PPR) between PFC and S1. In PFC paired pulse facilitation (PPF, 1.18, 0.98-1.52) prevailed associated with increased success rates in the 2nd pulse (1_x: 0.81, 0.73-0.91; x_1: 0.91, 0.82-0.97). In contrast, in S1 paired pulse depression (PPD, 0.78, 0.69-0.90) was found with high initial success rates (1_x: 1.00, 0.97-1.00; x_1: 0.99, 0.95-1.00). We observed the difference in short-term plasticity over a range of interstimulus intervals (ISIs) of 5 to 50 ms (SI Appendix, Figure S1A, B). Thus, the differences in synaptic efficacy and reliability between S1 and PFC during the first action potential correlate with differential short-term plasticity, with the emphasize in S1 being on the first transmission process while in PFC facilitation emphasized the subsequent release.

Next, we investigated transmission at synapses between L2/3 and L5PNs, using minimal extracellular stimulation near somata of L2/3PNs (Figure 1A). We found that also L2/3-L5PN synapses in PFC showed PPF (1.36, 1.14-1.88) as opposed to PPD in S1 (0.87, 0.72-0.99; Figure 1J, K). We observed this over a range of ISIs up to 100 ms (SI Appendix, Figure S1C, D). In addition, at PFC synapses the SD of the synaptic delays (0.46, 0.29-0.66) was again significantly larger than in S1 (0.25, 0.15-0.37; Figure 1L). Furthermore, we found larger EPSC decay time constants (τdecay) in PFC compared to S1, possibly reflecting a contribution of different postsynaptic glutamate receptors (SI Appendix, Figure S2A, B). Indeed, application of the N-methyl-D-aspartate receptor (NMDAR) antagonist 2-Amino-5-phosphonovaleriansäure (APV) removed the difference in τdecay between PFC and S1 (SI Appendix, Figure S2C-F). We extended the paired pulse experiments at L2/3-L5PN synapses in PFC to even older mice (P90-100) and to the mPFC (P21-26). We found that PPF persisted in the older age window (1.25, 1.14-1.41) and that also the synapses in mPFC showed PPF (1.14, 1.03-1.42; Figure 1K).

Together these results from glutamatergic synapses onto L5PNs show significant differences in the synaptic delays, efficacy, reliability and short-term plasticity between PFC and S1. The results would be consistent with area-specific differences in the functional coupling distances between VGCCs and SVs at the presynaptic active zones.

EGTA sensitivity of release is higher in PFC than in S1

To directly test for differences in the coupling distances, we investigated the sensitivity of release to synthetic Ca2+ chelators (Figure 2). The sensitivity of release to low millimolar concentrations of the kinetically slow Ca2+ chelator EGTA is a standard indicator of loose coupling (Eggermann et al., 2012). First, we performed recordings at pairs of L5PNs in mature PFC (Figure 2A, B). After 10 min of stable baseline recordings, the presynaptic neuron was patched a second time and equilibrated for 30 min with 10 mM EGTA-containing pipette solution in the whole-cell configuration. In these paired recordings we found a significant effect of EGTA onto release in PFC (0.67, 0.63-0.73). In control recordings a temporary initial increase in EPSC amplitudes after whole-cell access to the presynaptic neuron was evident. Although L5PNs do not express large concentrations of mobile buffers (Helmchen et al., 1996; Tran and Stricker, 2018; Bornschein et al., 2019b), their wash-out might have contributed to the temporary run-up. On the other hand, run-up correlated with a temporary decline in Rs in some recordings (3 out of 6). Run-up effects normalized after 20 min at the latest (1.02, 0.75-1.20). In identical experiments in S1, we had previously found release to be insensitive to 10 mM EGTA (1.04, 0.80-1.10; control: 1.08, 0.86-1.31)(Bornschein et al., 2019b). For comparison data from the previous study were reanalyzed for the test period between 20 and 30 min. EPSC amplitudes in S1 were sensitive to the same concentration of the kinetically fast Ca2+ chelator BAPTA (test period 10-20 min: 0.22, 0.13-0.33). These data provide evidence that the synapses between L5PNs in mature PFC do indeed operate with loose coupling, in contrast to their counterparts in S1.

Area-specific differences in EGTA sensitivity.

(A)Averaged, baseline-normalized EPSC amplitudes (means ± SEMs, 2-min bins) recorded from L5PN-L5PN pairs in PFC with standard pipette solution (control, orange) and after re-patching the presynaptic neuron with a pipette solution containing 10 mM EGTA (blue). Insets: Average EPSCs from one pair during baseline (-10 to 0 min) and test period (20 to 30 min) for control and EGTA (averages normalized to baseline). (B)Boxplot comparing the normalized EPSC amplitudes prior and following the application of EGTA and/or BAPTA with control recordings in L5PN pairs in PFC (*P=0.041, MWU) and S1 (***P<0.001, P=0.365, one-way ANOVA; Shapiro-Wilk normality test: P=0.597). S1 data are from ref. (Bornschein et al., 2019b). Note the significant EGTA effect. (C)As in (A) but for L2/3-L5PN connections in PFC. Following baseline recordings (10 min), slices were perfused for 30 min (solid line) with ACSF containing either 10 μM EGTA-AM (blue), 10 μM BAPTA-AM (red), or only the solvent DMSO/Pluronic (control, orange), and thereafter rinsed for 10 min with ACSF (test period). Insets: Averaged EPSCs during baseline and test period. (D)Summary of chelator-AM effects in PFC, mPFC and older PFC (P90-100; ***P<0.001, **P=0.009, *P=0.032, one-way ANOVA vs. control; Shapiro-Wilk normality test: P=0.666). Note the similar EGTA sensitivity of release in all PFC recordings (P=0.347, one-way ANOVA). (E)As in (C) but for L2/3-L5PN connections in S1 (control in light blue). (F)Summary of chelator-AM effects in S1 (**P=0.006, ANOVA on ranks). Note the absence of a significant EGTA effect in S1 (P=0.943, MWU).

We proceeded by comparing the EGTA sensitivity of release at L2/3-L5PN synapses between S1 and PFC, using extracellular stimulation and incubation with Ca2+ chelator-AM compounds (Figure 2C-F) (Baur et al., 2015; Kusch et al., 2018). Following 10 min of stable baseline recordings, slices were perfused with ACSF that either contained the solvent DMSO/Pluronic alone (control) or 10 µM of dissolved EGTA-AM or BAPTA-AM. The fast BAPTA interferes with release irrespective of the coupling topography, thereby, providing a positive control for the proper functioning of the AM-method (Eggermann et al., 2012). EPSC amplitudes were quantified during a subsequent 10 min period with perfusion with normal ACSF. During this test period, BAPTA significantly reduced EPSC amplitudes both in PFC (0.28, 0.12-0.32; control: 0.99, 0.84-1.02) and in S1 (0.08, 0-0.18; control: 0.85, 0.83-0.92). On the other hand, EGTA significantly reduced EPSC amplitudes only in PFC (0.54, 0.47-0.67; 55% of control) but not in S1 (0.85, 0.81-1.04; 100% of control). For a better comparison of EGTA effects some recordings were performed in PFC and S1 derived from the same animal to rule out interindividual effects (see example recording in Figure S3A-C).

In S1 we previously found that coupling switches from loose to tight during postnatal development between ∼P10 and ∼P20 (Bornschein et al., 2019b). To probe for a developmental retardation of this process in PFC, we extended the EGTA-AM experiments at L2/3-L5PN synapses to the older age-window of P90-100. Also, in these experiments we found a significant sensitivity of release to EGTA (0.76, 0.65-0.78; Figure 2D; SI Appendix, Figure S3D), indicating that loose coupling is a functionally persistent feature of synapses in PFC. Finally, we also extended the EGTA-AM experiments to synapses in mPFC and again found significant effects of EGTA on the EPSC amplitudes (0.65, 0.49-0.75; Figure 2D; SI Appendix, Figure S3E).

Together the results obtained with the slow Ca2+ chelator EGTA suggest that major glutamatergic synapses in mature PFC operate with loose coupling in contrast to tight coupling in the mature S1. The data further suggest that loose coupling is a developmentally persistent property of synapses in PFC. Finally, they indicate that the above functional differences between synapses in S1 and PFC originate from the differences in Ca2+ influx-release coupling.

Release probabilities are similar in PFC and S1

To test whether the release probabilities (pN) of the differentially coupled release sites in PFC and S1 differ, we performed multiple probability fluctuation analysis (MPFA) (Clements and Silver, 2000; Brachtendorf et al., 2025) at different extracellular Ca2+ concentrations ([Ca2+]e; Figure 3) (Bornschein et al., 2019b). The parabolic MPFA-fits to variance-mean plots from recordings of pairs of L5PNs in PFC yielded a median pN of 0.50 (0.37-0.55; Figure 3A, D). This value was slightly but not significantly smaller than pN of 0.66 (0.55-0.70), which we quantified previously for this connection in S1 (Bornschein et al., 2019b).

Quantal synaptic parameters in PFC and S1.

(A)MPFA of EPSC amplitudes recorded at the indicated [Ca2+]e from a pair of L5PNs in PFC. Top: EPSCs (gray, average in black) recorded in 1, 2 and 5 mM [Ca2+]e. Middle: Plot of EPSC amplitudes over time. Bottom: Corresponding mean-variance plot fitted with a parabola estimating the quantal parameters of release. pN is for 2 mM [Ca2+]e. (B)As in (A), but for a L2/3-L5PN connection in PFC. Recordings were made in the presence of 0.25 mM Kyn and 50 µM APV. (C)As in (B), but in S1. (D)Summary of quantal release parameters in L5PN pairs from PFC and S1 and in S1 with 1-2 mM γDGG (S1 data from (Bornschein et al., 2019b); P=0.121, *P=0.048, P=0.054, ANOVA on ranks). (E)Left: Example of evoked qEPSCs (top) and the corresponding amplitude histogram (bottom, gray range was used for calculating the qEPSC amplitude). Right: Summary of qEPSC amplitudes in PFC and S1 (**P=0.004, MWU). (F)Summary of pN values in L2/3-L5PN connections from PFC and S1 (P=0.485, MWU).

We also quantified pN at L2/3-L5PN synapses, both in PFC and in S1, using MPFA with extracellular stimulation (Baur et al., 2015). We found that pN was as high as at the L5PN-L5PN synapses in both areas with values (PFC: 0.60, 0.51-0.66; S1: 0.52, 0.40-0.64) being not significantly different between areas (Figure 3B, C, F). Interestingly, in our previous study (Bornschein et al., 2019b) we found that the developmental switch from loose to tight coupling at L5PN synapses in S1 also had no significant impact on pN.

From MPFA in paired recordings it is possible to estimate the number of release sites (N) and the quantal size (q), additionally. N was significantly smaller in PFC than in S1 being only 2.1 (1.9-2.3) rather than 8 (3-19). q tended to be smaller in PFC (6 pA, 5-7 pA) than in S1 (9 pA; 8-11 pA) and the difference became significant when evoked quantal EPSCs recorded in low [Ca2+]e were compared (PFC: 4 pA, 4-5 pA; S1: 9 pA, 7-11 pA). We consider it unlikely that postsynaptic receptor saturation or desensitization influenced the determination of quantal parameters since experiments with the competitive low-affinity glutamate receptor antagonist γ-DGG (γ-D-glutamylglycine) (Chanda and Xu-Friedman, 2010) performed at the S1 synapses had revealed no signs of this (pN=0.58, 0.52-0.60; N=8, 8-12; Figure 3D). Due to the smaller size of the EPSCs, testing on the PFC synapses was not possible. Estimating EPSC amplitudes (EPSC = N pN q; PFC, 6 pA; S1, 48 pA) and failure rates (F = (1-pN)^N; PFC, 0.25, S1, 0.0001) from the quantal parameters yielded values similar to those recorded (Figure 1). The determined parameters were therefore well suited to explain the area-specific differences in synaptic efficacy during individual action potentials.

Presynaptic Ca2+ signals are similar in PFC and S1

For a thorough interpretation of the above results knowledge about the presynaptic Ca2+ dynamics in the different synapses is required. We performed dual-dye two-photon Ca2+ imaging (Sabatini et al., 2002) to quantify volume-averaged Ca2+ signals at presumed presynaptic boutons located on axon collaterals of L5PNs in PFC and in S1. The imaged presynaptic boutons most likely connect to neighboring pyramidal cells, as connectivity between L5PNs in layer 5A is high (Feldmeyer, 2012). Nevertheless, a small proportion of other postsynaptic targets, such as interneurons, cannot be ruled out. We quantified single action potential-mediated elevations in green over red fluorescence signals (ΔG/R) that were converted to increases in intracellular calcium (Δ[Ca2+]i) (Figure 4) (Bornschein et al., 2019b). We found that neither the Ca2+ transients nor the decay time constants were significantly different between the cortical areas (Figure 4A-E; PFC: G/R=0.25, 0.23-0.31; Δ[Ca2+]i=279 nM, 244-453 nM; τdecay,1=9 ms, 6-19 ms; τdeay,2=91 ms, 75-120 ms; S1: G/R=0.29, 0.25-0.39; Δ[Ca2+]i=238 nM, 208-303 nM; τdecay,1=9 ms, 3-15 ms; τdecay,2=86 ms, 63-212 ms). The quantification of absolute basal [Ca2+]i was limited by the KD of Fluo5F (439 nM), which slightly underestimated basal [Ca2+]i values in comparison to quantification with OGB1 (KD=166 nM, Figure S4). Nevertheless, relative comparison of basal [Ca2+]i yielded no significant differences between PFC (30 nM, 21-34 nM) and S1 (22 nM, 13-38 nM; Figure 4F).

Presynaptic Ca2+ transients are similar in PFC and S1.

(A)Two-photon images of L5PNs in PFC (left) and S1 (right) filled with Fluo-5F and Alexa 594 (scale bar 50 µm). Insets: Magnifications showing presynaptic boutons from which Ca2+ transients are recorded (scale bar 5 µm). (B)G/R signals (top) evoked by single APs in individual cells (gray, average of 6-15 boutons each, grand average in black) and their conversion to Ca2+ increases (Δ[Ca2+]i, bottom). (C-F) Summary of G/R signals (C; P=0.112, MWU; numbers of boutons and cells in brackets), Δ[Ca2+]i (D; P=0.361), decay time constants τ1 and τ2 (E; P=0.532, 0.962), and basal Δ[Ca2+]i (F; P=0.597) in PFC and S1. Note, that there is no significant difference in presynaptic Ca2+ transients between PFC and S1.

The similarity in the presynaptic Ca2+ transients of boutons in PFC and S1 suggests that differences in the presynaptic Ca2+ influx are unlikely to be the cause for the differential Ca2+ chelator effects. Rather, they support the view that these differences result from loose coupling in PFC and from tight coupling in S1.

Estimate of the coupling distance in PFC verify loose microdomain coupling in PFC

To derive a quantitative picture of the coupling topography in L5PN synapses in PFC we used numerical computer simulations that were based on and constrained by the experimental data. First, we simulated the measured Ca2+ transient with the indicator dye present in the simulation (Figure 5A). Subsequently, the indicator dye was removed and a release sensor model for Synaptotagmin-1-triggered release from L5PN boutons was included in the simulations (Figure 5B, C; SI Appendix, Figure S5) (Bornschein et al., 2025). VGCCs were assumed to form a ring like structure around a vesicle (Figure 5D), similar to the release site topography in young S1 (Bornschein et al., 2019b). Release rates as reported by the sensor model were integrated over time to yield the pN values. The coupling distance between the sensor and the ring of VGCCs as well as the number of VGCCs forming the ring were iteratively varied until the simulation correctly predicted the experimental pN values from MPFA obtained under control conditions and in the presence of EGTA. By this procedure we quantified an average coupling distance of 48-51 nm between a microdomain of several VGCCs and the release sensor (Figure 5B; SI Appendix, Figure S5), similar to the estimate for these synapses in immature S1 (Bornschein et al., 2019b). Thus, a microdomain model similar to immature S1 is suitable to predict the data from mature PFC, whereas the same synapses in mature S1 operated with Ca2+ signaling nanodomains in which only 1-3 VGCCs at distances of 11-16 nm trigger fusion (Bornschein et al., 2019b; Bornschein et al., 2025).

Model predicts microdomain coupling at PFC synapses.

(A)Fit of the model (blue line) to the measured average Ca2+ signal (A; gray line) of L5PN boutons in PFC. The fit was obtained with the indicated number of open VGCCs. (B)pN values were simulated with the Syt1 priming model (Bornschein et al., 2019b; Bornschein et al., 2025) for control conditions and in the presence of EGTA with different VGCC numbers yielded different coupling distances (CD) in the model. The simulated pN were subtracted from and normalized to the experimental values, i.e. a value close to zero indicates best agreement between simulation and experiment. This was obtained with 9 open VGCCs at a CD of 51 nm (see C). (C)Simulated pN for 9 open VGCCs under control conditions at increasing CDs (blue line). The experimental pN value (median ± IQR, dashed line, gray range, Figure 3D) was reproduced at a CD of 51 nm (48-55 nm). (D)Illustration of the ring model used in B and C with a CD of 51 nm between 9 VGCCs (orange) and the release sensor of a vesicle (blue). (E)Examples of averaged normalized paired pulse recordings at 50 Hz at L2/3-L5PN synapses before (black) and after (orange) application of 5 µM Cd2+ in PFC. (F)Bar graph quantifying normalized PPRs after Cd2+ application in PFC and S1 (**P=0.005, P=0.667, paired t-Test; Shapiro-Wilk normality test: P=0.514). Note that Cd2+ increased the PPR in PFC.

To experimentally test whether overlapping Ca2+ domains from several VGCCs trigger release in PFC (microdomain), we analyzed the effects of the unspecific VGCC blocker Cd2+ onto the PPR. If fusion is triggered by such VGCC microdomains, application of a subsaturating concentration of Cd2+ will increase the PPR, while it will leave PPR unaffected if only a single or few VGCCs trigger release (Hefft et al., 2002; Bucurenciu et al., 2008; Scimemi and Diamond, 2012; Baur et al., 2015; Bornschein et al., 2019b). In paired pulse experiments, at L2/3-L5PN synapses in PFC we indeed found a significant increase in the PPR following the application of 5 µM Cd2+ (rPPR=1.20±0.06), whereas the average PPR remained unaffected in S1 (rPPR=1.01±0.10; Figure 5E, F). Thus, our data and simulations suggest that the nanotopography of PN synapses in mature PFC is reminiscent of the same synapses in young S1.

Discussion

Our results provide evidence that the same archetypes of glutamatergic synapses show area-specific differences in their functional release site nanoarchitectures in the mature neocortex. They suggest loose microdomain coupling as a synaptic correlate of higher order neocortical functions in PNs.

We did neither morphologically nor based on spiking patterns differentiate further between PN subtypes within a given layer. However, within a cortical area (S1 or PFC) we did not find a difference between glutamatergic synapses from L2/3 onto L5PNs and L5PN to L5PN synapses, neither with regard to the EGTA sensitivity of release nor with regard to pN. In particular, we found homogeneous results and similar variability in both, the examined connections in the PFC and in S1, with no discernible clustering in the data that would indicate stimulation of different cell populations. These findings suggest that excitatory inputs to L5PNs exhibit similar properties (PPR, pN, CD) irrespective of whether they originate in L2/3 or in neighboring PNs in L5A. However, we do see significant differences between synapses in the different cortical areas S1 and PFC. Thus, intra-area specific differences in morphology and spiking patters among PNs appear to be not reflected on the level of their synapses.

The differences in coupling gave rise to larger variability of synaptic delays (SDDelay) in PFC compared to S1 and were associated with differences in synaptic reliability and efficacy, whereas the pN values were not significantly different between areas. A major determinant of pN is the size of the Ca2+ signal at the release sensor. Accordingly, tightening of coupling can increase pN (Eggermann et al., 2012; Baur et al., 2015; Bornschein and Schmidt, 2019). However, large Ca2+ elevations at the release sensor can also be obtained with Ca2+ microdomains. At L5PN-L5PN synapses in S1 the developmental switch from microdomain to nanodomain coupling had no effect on pN (Bornschein et al., 2019b). At the calyx of Held (Iwasaki and Takahashi, 2001; Taschenberger et al., 2002; Fedchyshyn and Wang, 2005; Koike-Tani et al., 2008; Nakamura et al., 2015) and at cerebellar basket cell to Purkinje cell synapses (Chen et al., 2024) pN even decreased during development, whereas coupling got tightened. The decrease in pN was compensated by an increase in N at these synapses, which increased their reliability and efficacy during development. We found N to be larger in S1 than in PFC. Thus, it appears that differences in coupling are typically associated with other characteristic changes in the release parameters. Overall, this suggests that reliable CNS synapses switch from an initial state with high plasticity to a matured state with high reliability. On the other hand, PN synapses in the PFC appear to remain in a more juvenile state that favors plasticity over reliability.

Nanodomain coupling was also found in the peripheral nervous system, in particular at retinal (Singer and Diamond, 2003; Jarsky et al., 2010) and auditory (Moser and Beutner, 2000; Brandt et al., 2005) ribbon type synapses and at the neuromuscular junction (Harlow et al., 2001; Shahrezaei et al., 2006). These synapses have highly specialized properties and appear to be optimized for very reliable transmission and, in the case of ribbon synapses, also for high-frequency coding of sensory information (reviewed in Matthews and Fuchs, 2010; Eggermann et al., 2012). Thus, it appears that synapses in the sensory pathways, in particular those engaged in reliable high-frequency coding of sensory information, both in the periphery and in the lower processing stages of the CNS, up to primary sensory cortices, operate with nanodomain coupling. In the executing motor pathway, the neuromuscular junction uses nanodomain coupling and, as recent results from our group suggest, also PNs in the primary motor cortex (Yarim et al., 2026). It is tempting to speculate that complete loops from or to the primary cortices to their peripheral target organs operate with nanodomains. Microdomain coupling, on the other hand, appears to come into play only if integration of information from multiple sources and plasticity are the main focus, as at certain synapses in PFC (this study) or hippocampus (Vyleta and Jonas, 2014).

The microdomain was assumed to be formed by a ring-like structure of VGCCs around a vesicle (Figure 5D). This topography was chosen because such a microdomain was found to best predict the experimental data of transmitter release from PNs in young S1 (Bornschein et al., 2019b). Other previously described distributions of VGCCs suitable to reproduce release data cover random distributions of VGCCs (Scimemi and Diamond, 2012), VGCC clusters (Meinrenken et al., 2002; Nakamura et al., 2015; Rebola et al., 2019), and exclusion zones (Keller et al., 2015; Rebola et al., 2019). In the early S1, all of these models predicted a higher EGTA sensitivity of the microdomain, however, these models provided a poorer fit to the full set of the experimental data than the ring-like structure (Bornschein et al., 2019b). Since the experimental data from PNs in the mature PFC were similar to those in young S1, these other microdomain models were not tested explicitly here.

Differences in the sensitivity of release to low to moderate concentrations of EGTA (≤ 30 mM) are a standard indicator of differences in the coupling distance (e.g. Adler et al., 1991; Bucurenciu et al., 2008; reviewed in Eggermann et al., 2012; Vyleta and Jonas, 2014; Kusch et al., 2018; Bornschein et al., 2019b). pN is determined by the size of the Ca2+ signal at the release sensor and the binding kinetics and affinity of the sensor. The former in turn is determined by the details of the Ca2+ influx and the diffusional coupling distance between the VGCCs and the sensor. The similarity of Ca2+ signals between synapses in PFC and S1 (Figure 4) indicates that Ca2+ influx is similar between boutons, although more subtle differences in the influx kinetics may have remained undetected in these volume-averaged signals. Regarding sensor affinity, results in a previous study indicate that differences in EGTA sensitivity show differences in coupling rather than sensor affinity even if kon of the sensor and its affinity should differ as much as ten-fold, which appears to be an unlikely scenario given that even the two major isoforms of Synaptotagmin that trigger synchronous release differ by less than a factor of three to four in their affinity (Bollmann et al., 2000; Schneggenburger and Neher, 2000; Bornschein et al., 2025). Finally, the increase in the PPR induced by the application of Cd2+ further supports our conclusion of microdomain coupling in the PFC synapses (Scimemi and Diamond, 2012). Thus, although the absolute EGTA sensitivity is influenced by different factors, which necessitates data-constrained models for quantitative comparisons, the general sensitivity of release to EGTA indicates loose coupling.

The differences we observed in short-term plasticity suggest that S1 emphasized the first transmission process, whereas in PFC facilitation emphasized the subsequent release. The mechanisms of short-term plasticity are complex and include the regulation of pN, the speed of vesicle replenishment and the recruitment of release sites (Neher and Brose, 2018; Schmidt, 2019; Neher, 2023; Brachtendorf et al., 2025). Classically, pN was considered as the major determinant of short-term plasticity (e.g. reviewed in Zucker and Regehr, 2002; Feldmeyer and Radnikow, 2009). More recently other factors, including the number of occupied release sites, their replenishment or an increase in their occupancy, or the expression of endogenous Ca2+ buffers have been considered as more important determinants of short-term plasticity (Rozov et al., 2001; Blatow et al., 2003; Felmy et al., 2003; Matveev et al., 2004; Bornschein et al., 2013; Miki et al., 2016; Doussau et al., 2017; Jackman and Regehr, 2017; Neher and Brose, 2018). Irrespective of the detailed mechanism, we consider the differential emphasize on the available presynaptic efficacy between PFC and S1 as a signature of higher plasticity and flexibility in the processing of information in PFC compared to S1.

Short-term plasticity changes during postnatal development at different cortical PN connections without alterations in pN (Reyes and Sakmann, 1999; Bornschein et al., 2019a). For L5PN to L5PN connections these differences were found to result from the maturation of an intermediate replenishment vesicle pool (Bornschein et al., 2019a). Such pool maturation may also underlie the elimination of layer-specific differences in short-term plasticity between L2/3-L5B and L5B-L5B synaptic connections that were evident in young rats but eliminated during the first weeks of postnatal development (Reyes and Sakmann, 1999). Since postnatal development follows the same time-course in mouse PFC and S1 and occurs predominantly within the first two weeks after birth, significant layer-specific differences in PN synapses are unlikely in both areas in our experimental time window (Kroon et al., 2019). Consistently, we found similar PPRs at L2/3-L5PN synapses and L5PN-L5PN synapses in both areas, with facilitation in PFC and depression in S1, irrespective of the presynaptic PN synapse type.

Two other previous studies on L5APN (Frick et al., 2008) and L5BPN (Markram et al., 1997) connections concluded that these synapses operate with high release probability, which nicely agrees with our previous (Bornschein et al., 2019b) and current results. It is remarkable that we did not even detect any differences between the L2/3-L5PN and L5PN-L5PN synapses within a given cortical area. Overall these results from different studies (Markram et al., 1997; Reyes and Sakmann, 1999; Frick et al., 2008; Bornschein et al., 2019a; Bornschein et al., 2019b) may indicate that variability on the synaptic level between PNs of a given area is not pronounced. In order to further substantiate this, we determined the relative variability in median EPSC amplitudes to test whether there is a higher variability of recorded cell types in PFC compared to S1. The relative MAD (median absolute deviation) of EPSC amplitudes was 0.46 in PFC and 0.50 in S1. The similarity in these values argues against higher cell-type variability in PFC compared to S1.

Our results have implications for long-term plasticity at neocortical PN synapses. Both, long-term potentiation (LTP) and depression (LTD) have a presynaptic locus at these synapses and likely involve changes in pN, but the final steps that alter release are not well understood (Feldman, 2009; Castillo, 2012; Feldman, 2012). A recent study on L5PN synapses in S1 showed an increase in the number of primed vesicles as a major mechanism of LTP (Weichard et al., 2023). PN synapses in S1 are already tightly coupled and therefore a further decrease in the coupling distance is an unlikely mechanism of LTP. However, this does not exclude that in S1 LTD induction breaks tight coupling and vice versa that LTP in PFC involves tightening of coupling. In addition, in PFC we found an increased contribution of NMDAR on the postsynaptic site. Thus, these synapses have increased potential for plasticity on the pre- and the postsynaptic site. During induction of long-term plasticity, the postsynaptic site signals back to the presynaptic site via retrograde messengers (Regehr et al., 2009) and loose coupling will provide the substrate for enhanced regulatory capacity (Vyleta and Jonas, 2014).

In general, loose coupling provides more flexibility to regulate [Ca2+]i at the release sensor (Vyleta and Jonas, 2014). Prefrontal networks carry out higher-order computations that transform sensory information and memory to perform flexible output such as delayed response, inference and planning (Narayanan et al., 2025). Our findings suggest that loose coupling provides a synaptic foundation for such flexible neocortical network functions.

Methods

Slice preparation

C57BL/6J mice at P21-26 of either sex were used in this study. All experiments were performed in accordance with the guidelines for the welfare of experimental animals issued by the European Communities Council Directive (2010/63/EU) and with the German Protection of Animals Act (Tierschutzgesetz). Mice were bred in the animal facility of the Medical Faculty of Leipzig University and were housed in individually ventilated cages in a specific pathogen free environment and in a 12h/12h light dark cycle with access to food and water ad libitum. Experiments were approved by the animal welfare office of the University Medical Center, Leipzig, as well as the local governmental authorities (Landesdirektion Leipzig, registration numbers T10/20, T05/21-MEZ).

C57BL/6J mice at P21-26 and P90-100 of either sex were decapitated under deep Isoflurane (Curamed) inhalation anesthesia. The brain was excised rapidly and placed in cooled (0-4°C) artificial cerebrospinal fluid (ACSF) containing (in mM): 125 NaCl, 2.5 KCl, 1.25 NaH2PO4, 26 NaHCO3, 1 MgCl2, 2 CaCl2, and 20 glucose, equilibrated with 95% O2 and 5% CO2 (pH 7.3-7.4). Coronar neocortical slices (150-250 μm thick) were cut from the lateral PFC, medial PFC (mPFC) or S1 region (Figure 1A) with a vibratome (HM 650 V, Microm). For some paired recordings between L5PNs and for calcium imaging experiments parasagittal slices (200 µm) were prepared from PFC region. Slices were incubated for 30 min at 35°C and subsequently stored at room temperature. For experiments, slices were transferred to a recording chamber and continuously perfused with ACSF (2-3 ml per min, supplemented with 10 μM (-)-bicuculline methiodide (Tocris) at 31-33°C. Unless stated otherwise, chemicals were from Sigma-Aldrich.

Electrophysiological recordings

Patch-clamp recordings from L5PNs located in the upper layer 5 (L5A in S1) were established according to the criteria described in detail in our previous work on this connection in S1 (Bornschein et al., 2019b; Bornschein et al., 2025). Presynaptic neurons were stimulated extracellularly in upper layer 2/3 (L2/3-L5PN connections) straight above the patched L5PN or in on-cell mode in L5A right next to the postsynaptic cell (L5PN-L5PN connections; Figure 1).

Patch pipettes were prepared from borosilicate glass (Hilgenberg) with a PC-10 puller (Narishige) and had final resistances of 6-8 MΩ when filled with the following standard pipette solution (in mM): 150 K-gluconate, 4 NaCl, 3 MgCl2, 3 Na2ATP, 0.3 NaGTP, 0.05 EGTA, 10 KHEPES, dissolved in purified water. The pH was adjusted to 7.3 with KOH. Recordings were performed under optical control (BX51WI, Olympus), using an EPC10/2 amplifier and Patchmaster software (version v2x90.2, HEKA). EPSCs were recorded in the whole-cell configuration at a holding potential (Vhold) of -90 mV (online corrected for a liquid junction potential of 16 mV), filtered at 5 kHz and sampled at 10 kHz. Series resistance (Rs) was continuously compensated to a fixed value between 10 and 15 MΩ. The average uncompensated Rs was 18±1 MΩ in L5PN-L5PN paired recordings (n=41 cells, mean ± SEM) and 23±1 MΩ in L2/3-L5PN recordings (n=133 cells). Holding current (Ihold) was also monitored continuously and was <180 pA (-134±16 pA in L5PN-L5PNs; -168±11 in L2/3-L5PNs). Typically, recordings were excluded if Rs exceeded 30 MΩ and Ihold fell below -300 pA. In paired recordings, presynaptic L5PNs were stimulated in on-cell configuration (200-500 mV, 1-2 ms).

In the chelator wash-in experiments, presynaptic neurons were repatched with a pipette solution supplemented with 10 mM EGTA (K-gluconate concentration was reduced to 135 mM to adjust osmolarity) and whole-cell configuration was established to allow EGTA perfusion of the presynaptic neuron. EPSCs were evoked every 10 s for at least 30 min (Figure 2, S3). Averaged EPSC amplitudes were calculated before (≥10 min baseline) and 20-30 min after repatching (test period). For recordings from L2/3-L5PN connections ACSF-filled patch pipettes were placed close to the somata of PNs in L2/3 and extracellular stimulation was performed using an ISO-Stim 01 DPI (NPI electronics). Minimal stimulation (0.5 to 5 V) was used and yielded EPSC amplitudes of ∼40 pA (47±3 pA, n=131 pairs), making it likely that we investigated transmission mainly between single or few L2/3PNs and L5PNs in these experiments. Following 10 min of stable recordings in standard ACSF (baseline), the bath solution was exchanged for ACSF containing either 0.1% DMSO and 0.01% Pluronic (control solution), 10 µM EGTA-AM or 10 µM BAPTA-AM in DMSO/Pluronic for 30 min (incubation period) and thereafter replaced by standard ACSF again for another 10 min (test period). Averaged EPSC amplitudes were calculated during baseline and test period. The time courses of chelator effects on EPSC amplitudes were analyzed by binning EPSC amplitudes within 2 min intervals of recording time to average amplitudes. Chelator effects were expressed as averaged EPSC amplitudes during test period normalized to the corresponding baseline values.

VGCCs were inhibited by bath-application of ACSF containing a subsaturating concentration of 5 µM CdCl2 (Figure 5E, F). NMDA receptors were blocked by the selective antagonist 2-Amino-5-phosphonovaleriansäure (APV, 50 μM; SI Appendix, Figure S2). For assessing the effects of the individual blockers, EPSCs were recorded in ACSF for at least 10 min (10 s intervals), subsequently the blocker was perfused and the blocker effects were quantified after a stable block had been established (at least 5 min test period).

Quantification of quantal synaptic parameters

Quantal synaptic parameters were estimated from parabolic fits to mean-variance (MV) plots of EPSC amplitudes recorded at different [Ca2+]e (1, 2, and 5 mM, ≥30 repetitions per concentration; [Mg2+]e was correspondingly adjusted to 2, 1, 0 mM, respectively), assuming binominal release statistics (Clements and Silver, 2000; Scheuss et al., 2002; Silver, 2003)(Figure 3). The recordings always started in 2 mM [Ca2+]e. The variance of EPSCs (σ2) was calculated according to ref. (Scheuss et al., 2002), plotted against the averaged amplitudes (I), and fitted by a parabola of the form: (1) σ2=Iq−I2N(1+CVII2)+qICVI2 where q is the quantal size, N a binominal parameter, and CVI and CVII the coefficients of intrasite and intersite quantal variability, assumed to be 0.3 (Clements and Silver, 2000). For CV assumptions, that deviate from the standard value of 0.3, deviations in the calculated pN values of less than 10% are to be expected (Schmidt et al., 2013; Bornschein et al., 2019b). The fits were constrained to pass through zero. The variance of the variance was calculated according to ref. (Meyer et al., 2001). Stationarity of EPSC amplitudes was established when, after full exchange of the bathing solutions (≥5 min), EPSC amplitudes no longer showed a tendency to increase or decline. For MV analysis in L2/3-L5PN connections, the bath solution was supplemented with the competitive AMPA receptor antagonist kynurenic acid (Kyn, 0.25 mM), which relieves their desensitization and saturation (Neher and Sakaba, 2001) and with 50 µM APV to prevent NMDAR activation. Due to the small size of EPSCs in L5PN-L5PN connections no blocker was used in these recordings. The size of qEPSCs was quantified from evoked miniature EPSCs recorded in low [Ca2+]e (1 mM) and high [Mg2+]e (2 mM).

Ca2+ imaging

Action potential-evoked (current injection of 1-4 nA for 1-2 ms) fluorescence changes were recorded at boutons located on axon collaterals of L5PNs as described previously (Bornschein et al., 2019b; Bornschein et al., 2025). L5PNs were filled with EGTA-free, Fluo-5F (200 µM, Invitrogen) and Alexa-594 (50 µM, Molecular Probes) containing pipette solution via somatic whole-cell patch-pipettes (Figure 4). Basal [Ca2+]i values in S1 were additionally determined by OGB-1 (100 µM, Invitrogen; Figure S4). Cells were dialyzed for at least 20 min to yield sufficient equilibration with the dyes. Volume averaged fluorescence signals were recorded in point mode at 500 kHz temporal resolution and subsequently binned to 500 Hz, using a custom-build two-photon microscope based on a Fluoview-300 scanner (Olympus), a 60x/0.9 N.A. objective, a mode-locked Ti:sapphire laser (Tsunami, Newport-Spectra Physics, set to a center wavelength of 810 nm), and a Pockels cell (350-80 KD*P, Conoptics). The fluorescence background was determined in a subsequent point mode recording performed near the boutons under investigation. On average 11 (but at least 3) boutons per L5PN were recorded and averages per cell were calculated to include a large number of putative presynapses. The fluorescence signals were filtered (HC647/75, Semrock HC525/50, 720-SP, AHF), detected by two external PMT modules (H7422-40, Hamamatsu; PMT-02M/PMM-03, NPI electronics) monitoring red and green epifluorescence, respectively at fixed PMT voltages, and digitized with the Fluoview system. The Ca2+-dependent green fluorescence was normalized to the Ca2+-insensitive red fluorescence and expressed as background-corrected ΔG/R (Sabatini et al., 2002). ΔG/R signals were converted to changes in [Ca2+]I based on an in vitro quantification of the KD of Fluo-5F (439 nM) and OGB-1 (166 nM), respectively, in our pipette solution, adjusted with CaEGTA and K2EGTA (100 mM stock solutions, 10 mM HEPES added) to contain different free Ca2+ concentrations that were calculated with the MaxChelator (https://somapp.ucdmc.ucdavis.edu/pharmacology/bers/maxchelator/CaMgATPEGTA-NIST.htm). Rmin and Rmax values were measured after each successful recording in sealed patch pipettes that were placed in the slices near the recorded cells. The pipettes contained the EGTA-free, Fluo-5F- and Alexa-594-containing pipette solution with either 0 mM CaCl2 and 10 mM K2EGTA (zero Ca2+ solution) or 20 mM CaCl2 (high Ca2+ solution).

Modeling

The kinetic gating scheme of the VGCCs and the reaction schemes for Ca2+ binding to buffers and transmitter release were converted to ordinary differential equations (ODEs). The system of ODEs was numerically solved using “NDSolve” or “NDSolveValue” of Mathematica 14 (Wolfram) as described previously (Schmidt et al., 2013; Bornschein et al., 2019b; Bornschein et al., 2025). Spatial resolution was achieved by placing the ODEs in concentric hemi-shells (1 or 2 nm thickness) that were coupled via diffusion.

The presynaptic action potential was modeled with a half-duration of 0.2 ms, τrise of 0.1 and τdecay of 0.2 ms (Borst et al., 1995). Voltage dependent activation of VGCCs was simulated using the gating scheme for P/Q-type channels, a single channel conductance of 2.5 pS, and a Ca2+ equilibrium potential of 130 mV (Li et al., 2007). The model included ATP as mobile buffer and immobile endogenous buffers with a buffer capacity (κE) of 25 (Tran and Stricker, 2018). Model parameters are given in Table S1 (SI Appendix).

The model for each channel consisted of five closed (C) and one open (O) state. The transition between the first five steps was voltage dependent and the last step voltage independent: (2) C0⇄α1(V)β1(V)C1⇄α2(V)β2(V)C2⇄α3(V)β3(V)C3⇄α4(V)β4(V)C4⇄ab0 a and b are the rate constants for transitions between C4 and O, and αi(V) and βi(V) are the voltage dependent forward and backward transition rates that are given by: (3a) αi(V)=αi,0exp⁡(V/ki)i=1−4 (3b) βi(V)=βi,0exp⁡(−V/ki)i=1−4 where αi,0 and βi,0 are the forward and backward rate constants at 0 mV and ki a slope factor.

Ca2+ binding to all buffers (B) was simulated by second order kinetics with forward and backward binding rate constants kon and koff, respectively: (4a) d[Ca]dt=−kon[Ca][B]+koff[CaB] (4b) d[B]dt=−d[CaB]dt Diffusion of all species (X) was simulated according to (5) d[X]dt=DxAn,n±1Vnr([X]n±1−[X]n) where DX is the diffusion coefficient, A the surface, V the volume and r the radius of a shell (n). Ca2+ was cleared by a linear, surface-based extrusion mechanism driven by the difference between [Ca2+]i and the resting [Ca2+]i (30 nM according to the imaging data).

For fitting the model to the average experimental Ca2+ transient, the number of VGCCs was adjusted, yielding 33 VGCCs (Figure 5). The other free parameter of the simulation was the extrusion rate, which however affected only the later phase of the decay of the transient. During fitting, the model included the Ca2+ indicator dye. Fitting was first performed in a single compartment model (Helmchen and Tank, 2005). Subsequently, fitting was performed in the spatially resolved model with the numbers of VGCCs remaining unaltered and [Ca2+]i as reported by the dye was calculated from the concentration of the Ca2+-dye complex assuming equilibrium conditions (Schmidt et al., 2003). Since L5PNs are characterized by an immobile and small κE with unknown binding kinetics, a slow (EBs) and a fast immobile endogenous buffer (EBf) were assumed (Helmchen et al., 1996; Ohana and Sakmann, 1998; Tran and Stricker, 2018). The fractions of EBs and EBf were adjusted to optimize the fit to the measured Ca2+ transient and to sum up to κE of 25. Following fitting, the indicator dye was removed from the model and the recently developed model for release from L5PN terminals was included (Bornschein et al., 2025). Either the “Syt1 priming model” (Equation 6; Figure 5) or the “Syt1 model” (Equation 7; SI Appendix, Figure S5) were placed at varying distances from the Ca2+ sources. In the priming model (Equation 6), a single Ca2+-dependent priming step with Michaelis-Menten kinetics (KM) and a priming rate kprim is assumed. V is the vesicular release sensor of a primed vesicle (Equation 6, 7) that binds a maximum of 5 Ca2+ ions. kon and koff are the forward and backward Ca2+ binding rate constants, βi (i=0-4) is a cooperativity factor, and γ the fusion rate. pN was obtained by integrating the release rates over time.

A ring like arrangement of the VGCCs around a vesicle was assumed (Figure 5D; ring 1 model in ref. (Bornschein et al., 2019b)). The number of VGCCs driving release was continuously reduced from the maximal value of 33 (Figure 5A) and the coupling distances at which the simulated pN values matched the experimentally determined pN were obtained. Subsequently, EGTA was included and the consistency between the experimentally quantified EGTA effect and the simulated EGTA effect at a given coupling distance was determined (Figure 5B). The best match between experiments and simulation was obtained for 9 VGCCs at a coupling distance of 51 nm (Figure 5C).

Quantification and statistical analysis

Data from L5PN-L5PN pairs in PFC were compared to previously published data from these pairs in S1 (Table 1)(Bornschein et al., 2019a; Bornschein et al., 2019b). Synaptic responses were determined by fitting a product of two exponential functions to the baseline-subtracted currents, which allows for independent adjustment of the time constants of the rising and falling phases and minimizes noise effects (Bornschein et al., 2013). Synaptic delays were determined from the onset of stimulation to the fitted onset of the EPSC. PPRs were calculated by dividing the second amplitude of two consecutive EPSCs by the first.

Electrophysiological data and Ca2+ imaging data were analyzed using custom written routines in Igor Pro (version 6.32 and 8.03, Wavemetrics). Data are shown as median and interquartile range (IQR) or as mean ± standard error (SEM). Summarized data are either shown in boxplots as median ± IQR with mean values included as dashed lines in addition or in bar graphs as mean ± SEM. Normality was tested using the Shapiro-Wilk test. Normally distributed data were compared with the t-test (two groups) or a one-way ANOVA (more than two groups). Non-normally distributed or small samples of data were compared with the Mann-Whitney-U rank sum test (MWU; two groups) or a Kruskal-Wallis ANOVA on ranks (more than two groups). For multiple comparisons post-hoc testing was performed with the Holm-Sidak (one-way ANOVA) or Dunn’s method (ANOVA on ranks). To compare pre- and post-treatment data the paired t-test or the Wilcoxon signed rank test (WSR) was used, depending on the distribution of the data. All statistical tests were two-tailed. P values are indicated as *P<0.05, **P< 0.01 and ***P<0.001. The number of experiments n represents the number of cell pairs, cells or boutons and was chosen sufficiently high to permit reliable statistical analysis. All of the statistical details of experiments can be found in the figures and corresponding figure legends. Statistics were performed with Sigma Plot 11.0 (Dundas Software) and Mathematica 14.